---
title: "Analytical Confidence Intervals for State Occupancy Probabilities"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Analytical Confidence Intervals for State Occupancy Probabilities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

State occupancy probabilities (SOPs), the probabilities of being in each state
at each time, are estimated from a fitted transition model, so they inherit
its uncertainty. `inferences(method = "delta")` quantifies this uncertainty
analytically. It calculates how each SOP changes when the model coefficients
change slightly and combines these sensitivities with the uncertainty in the
coefficients (the *delta method*). Unlike simulation or the bootstrap, it
uses no random draws, so it is fast and gives the same answer every time.

This article explains

- how to choose between conditional and unconditional variance;
- how the two variances are calculated;
- which patients enter the calculation, and why patient IDs matter;
- how the confidence limits are formed;
- how to extract variances and covariances; and
- which models and comparisons are supported.

Note that `method = "delta"` must be requested explicitly. The default,
`method = "mvn"`, simulates model coefficients instead.

## Fit a model

We simulate 800 patients over 30 days with `sim_actt2_markov()`, which uses
the eight-state NIAID scale described in `vignette("full-po-sops")`, and fit
a proportional-odds transition model with `orm_markov()`. `id_var = "id"`
identifies the rows that belong to the same patient, which the variance
calculations need.

```{r setup}
library(mostr)
library(rms)
library(patchwork)
library(ggplot2)

N_PATIENTS <- 800
FOLLOW_UP <- 30
SEED <- 20260718
ABSORB <- "8"

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_prob = 0.5,
  absorbing_states = as.integer(ABSORB),
  seed = SEED,
  treatment_effect = -0.5
)

plot_sops(raw_data, linetype_var = "tx", facet_var = NULL)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB)
)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit <- orm_markov(
  y ~ tx + time + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)
```

## Choose conditional or unconditional variance

An averaged SOP, such as the probability of being at home on day 14 if every
patient had received treatment, depends on two things: the model coefficients
and the patients over whom we average. The `vcov` argument of `inferences()`
decides which of these count as uncertain.

- **Conditional variance**, `vcov = "conditional"`, accounts for uncertainty
  in the estimated coefficients but treats the patients' starting states and
  covariates as given. It answers: how precisely have we estimated the
  average *for these particular patients*?
- **Unconditional variance**, `vcov = "unconditional"`, additionally accounts
  for which patients happened to be sampled, and for the fact that the same
  patients were used to estimate the coefficients. It answers: how precisely
  have we estimated the average *in the population the patients were drawn
  from*?

Both give the same point estimates; only the standard errors and confidence
intervals differ. Unconditional intervals are usually wider, most visibly
early in follow-up, when SOPs depend strongly on the patients' starting
states.

Averaged SOPs from `avg_sops()` and comparisons from `avg_comparisons()` use
unconditional variance by default. Patient-level SOPs from `sops()` support
only conditional variance, since there is no average over sampled patients.

## Conditional variance

Let $i = 1,\ldots,n$ index
patients, and let $\beta$ be the vector of all $p$ model coefficients, with
estimate $\hat\beta$. For patient $i$, the column vector $h_i(\beta)$ stacks
the predicted probabilities $S_{it}(l)$ for all requested times $t$ and
states $l$. Write $h_i = h_i(\hat\beta)$ for the fitted values and
$\bar h = n^{-1}\sum_{i=1}^n h_i$ for their average, which is what
`avg_sops()` reports.

The matrix $G_i$ describes how patient $i$'s predictions change with the
coefficients: it contains the derivatives of $h_i(\beta)$ with respect to
$\beta^\top$ at $\hat\beta$, with one row per prediction and one column per
coefficient. The average prediction changes according to
$\bar G = n^{-1}\sum_{i=1}^n G_i$. If $V_\beta$ is the estimated covariance
matrix of the coefficients, the conditional covariance matrix of the averaged
SOPs is

$$
\bar G V_\beta \bar G^\top.
$$

The square roots of its diagonal entries are the standard errors.

By default, $V_\beta$ is the fit's *patient-cluster robust* covariance, which
allows for the correlation between a patient's repeated transitions. There are two
fitting options in `orm_markov()` and `vglm_markov()` that allow adjustment for small
samples:

- `type = "HC1"` multiplies it by $(m - 1)/(m - p)$, where $m$ is the number
  of rows used in fitting (rows with positive weight, for `orm_markov()`) and
  $p$ is the number of coefficients. `type = "HC0"` applies no such factor, and is used by default.
- `cadjust = TRUE` multiplies it by $n/(n - 1)$, where $n$ is the number of
  patients. Fits with `id_var` use this adjustment by default.

Conditional variance, like MVN simulation, also works for weighted and
penalized `orm_markov()` fits whose fitted covariance is valid.

```{r conditional-variance}
# compute sops
t1 <- Sys.time()
avg_conditional <- avg_sops(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |> inferences(
  method = "delta",
  vcov = "conditional",
  conf_type = "auto"
)
t2 <- Sys.time()
t2 - t1

# compute time in state
time_conditional <- avg_time(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    vcov = "conditional", 
  )

# compute differences in time in state
comp_conditional <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB,
  estimand = "time_in_state",
  comparison = "difference"
) |>
  inferences(
    method = "delta",
    vcov = "conditional", 
    # specify null hypothesis, here no difference on additive scale
    null = 0
  )

p1.sops <- plot_sops(avg_conditional, linetype_var = "tx", facet_var = NULL) +
  labs(title = "Conditional variance")

p1.c <- plot_comparisons(comp_conditional) +
  labs(title = "Conditional variance")
```

## Unconditional variance

The unconditional calculation adds a second source of variation: had a
different set of patients enrolled, both the average of their predictions and
the fitted coefficients would have changed, and these two changes are
related. Each patient's contribution to both is captured in one vector,

$$
\phi_i = h_i - \bar h + \bar G A^{-1}s_i.
$$

The first part, $h_i - \bar h$, is how far patient $i$'s predictions lie from
the average. The second part is the patient's influence on the coefficients,
translated into its effect on the average prediction through $\bar G$. In it,

- $s_i = \sum_{r \in \mathcal{I}_i}\psi_r$ is the patient's *score*: the sum,
  over the patient's fitting rows $\mathcal{I}_i$, of the derivatives
  $\psi_r = \partial\ell_r/\partial\beta$ of each row's log-likelihood
  $\ell_r$, evaluated at $\hat\beta$;
- $A^{-1} = nB$, where $B$ is the inverse of the fit's total information
  matrix, before any `type` or `cadjust` correction. With the default
  observed information, $A$ is the negative derivative of the average
  patient score; a fit that uses a different (working) information matrix
  uses that matrix instead.

Thus $A^{-1}s_i$ approximates how much patient $i$ moves the estimated
coefficients. The unconditional covariance matrix is the sample covariance of
the $\phi_i$ divided by $n$:

$$
V_{\mathrm{unconditional}} =
  \frac{1}{n(n-1)} \sum_{i=1}^n
    (\phi_i - \bar\phi)(\phi_i - \bar\phi)^\top,
\qquad \bar\phi = n^{-1}\sum_{i=1}^n \phi_i.
$$

In R terms, this is `stats::cov(phi) / n`, where `stats::cov()` uses the
usual divisor $n - 1$. Because the two parts of $\phi_i$ are combined before
taking the covariance, their correlation in the observed sample is retained.
The fitting options `type` and `cadjust` do not change this estimate.

```{r unconditional-variance}
avg_unconditional <- avg_sops(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |> inferences(
  method = "delta",
  vcov = "unconditional",
  conf_type = "wald"
)

# compute time in state
time_unconditional <- avg_time(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    vcov = "unconditional", 
  )


comp_unconditional <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB,
  estimand = "time_in_state",
  comparison = "difference"
) |>
  inferences(
    method = "delta",
    vcov = "unconditional",
    null = 0
  )

p2.sops <- plot_sops(avg_unconditional, linetype_var = "tx", facet_var = NULL) +
  labs(title = "Unconditional variance")

p2.c <- plot_comparisons(comp_unconditional) +
  labs(title = "Unconditional variance")
```

The plots below compare the two choices, first for the SOPs and then for the
differences in expected time in each state (treated minus control). The
point estimates are identical; the intervals differ.

```{r compare-variances}
p1.sops + p2.sops + plot_layout(guides = "collect")
p1.c + p2.c
```

We can also plot the expect time in each state by arm using Grotta bars:

```{r}
plot_time(time_conditional, show_labels = FALSE)
```

If we want to show uncertainty, we can plot the expected time differently as well:

```{r}
plot_time(time_conditional, show_labels = FALSE, type = "pointrange")
```

### Requirements for unconditional variance

Unconditional variance needs each patient's scores, so it is available only
when

- the model was fitted with `orm_markov()` or `vglm_markov()`, which store
  the patient information needed; models fitted directly with `rms::orm()` or
  `VGAM::vglm()` cannot be used for analytical intervals at all;
- the average is taken over the patients stored with the model, not over
  user-supplied `newdata`; and
- the `orm_markov()` fit is neither weighted nor penalized.

If a requirement is not met, `inferences()` stops with an error rather than
silently switching to conditional variance. With `newdata`, for example,
request `vcov = "conditional"` explicitly.

## Which patients are included in the average?

Without `newdata`, predictions start from each fitted patient's row at the
first follow-up time, `first_followup_time`: that row supplies the patient's
starting state (`yprev`) and covariates. The model stores these rows before
dropping rows with missing outcomes. Therefore:

- each patient must have exactly one row at that time, with ID, predictors,
  time, and `yprev` all observed;
- the outcome on that row may be missing, provided the patient contributes at
  least one other transition to the fit;
- a later row cannot stand in for missing starting information.

For numeric time, `first_followup_time` defaults to 1, which must be the
earliest scheduled time. If you set another value, it must still be the
earliest. For factor or character time, you must set it explicitly; see
`vignette("factor-time-orm-sops")`.

Neither variance corrects bias caused by missing observations whose
missingness depends on unobserved outcomes.

## Why patient IDs matter

Each patient contributes many transitions to the model, and these are not
independent. Treating them as independent would typically make standard
errors too small. mostr therefore groups the fitting rows by patient,
identifying patients as follows:

1. If you pass `cluster` to `inferences()`, its entries identify the patient
   for each fitting row, in the order of the fitting data.
2. Otherwise, mostr uses the `id_var` column given to `orm_markov()` or
   `vglm_markov()`.
3. Without patient IDs, patient-cluster robust variance cannot be
   calculated.

For conditional variance, you can instead supply your own coefficient
covariance matrix through `vcov`. It must cover all model coefficients, with
matching names. Supplying a matrix always selects conditional variance,
because a matrix alone does not carry the patient-level scores that
unconditional variance needs.

Grouping by patient corrects the standard errors for within-patient
correlation. It does not correct for a transition model that describes the
data poorly, nor for bias from missing observations.

## Logit and Wald intervals

`conf_type` controls how confidence limits are formed from the estimate and
its standard error.

- `conf_type = "auto"`, the default, forms intervals for probabilities on
  the logit scale and transforms the limits back. This keeps them between 0
  and 1.
- `conf_type = "wald"` uses the estimate plus or minus a normal quantile
  times the standard error. Its limits can fall below 0 or above 1.

Differences between arms can be negative, so analytical intervals for
differences always use the Wald form; `conf_type = "auto"` selects it
automatically.

If an estimated probability is exactly 0 or 1, its logit is undefined and the
logit-scale limits are returned as `NA` with a warning, even when the
standard error is 0. Such values can arise from rounding and do not
necessarily mean that the outcome is certain. This applies to absorbing
states too: patients who enter one stay there, but the probability of
entering it is still estimated.

Each interval covers one estimate at the stated confidence level. The level
does not apply jointly to all intervals.

The example below compares the average probability of being outside hospital
(states 1 and 2 combined) between arms, as treated minus control.

```{r comparison}
sop_difference <- avg_comparisons(
  fit,
  variables = list(tx = c(0, 1)),
  estimand = "sop",
  state_sets = c("1", "2"),
  comparison = "difference",
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    vcov = "conditional",
    conf_type = "auto"
  )

head(sop_difference)
```

The standard error of a difference accounts for both arms' predictions coming
from the same fitted model and the same patients; the two arms are not
treated as independent.

`estimand = "time_in_state"` compares the expected time spent in a state
instead, as in the plots above. For visits recorded as a factor, supply the
elapsed time of each visit; `vignette("factor-time-orm-sops")` shows how, and
how to include the observed baseline state.

## Extract variances and covariances

The output of `inferences()` already contains standard errors and confidence
limits. To learn how the uncertainty in different estimates is related, for
example to test a combination of estimates yourself, use `vcov()`. It
returns the covariance matrix of the estimates: its diagonal entries are
variances, and the off-diagonal entries are covariances between pairs of
estimates. The matrix reflects the conditional or unconditional choice made
in `inferences()`.

The example requests the first six estimates only. The square roots of the
diagonal are their standard errors, matching the `std.error` column.

```{r accessors}
rows <- seq_len(min(6L, nrow(avg_conditional)))

conditional_covariance <- stats::vcov(avg_conditional, rows = rows)
unconditional_covariance <- stats::vcov(avg_unconditional, rows = rows)

sqrt(diag(conditional_covariance))
sqrt(diag(unconditional_covariance))
```

Without `rows`, `vcov()` returns a covariance for every pair of estimates.
This can use a lot of memory, because doubling the number of estimates
quadruples the number of entries, so request only the rows you need. The
option `mostr.delta_max_bytes`, 256 MiB by default, caps the largest memory
allocations made by analytical inference and by `vcov()`.

If you want to see the derivatives $\bar G$ themselves, the internal function
below returns them for the selected estimates: one row per estimate and one
column per coefficient. They enter both variances, but they do not include
the patient-sampling part of the unconditional variance.

```{r coefficient-derivatives}
coefficient_derivatives <- mostr:::get_jacobian(avg_conditional, rows = rows)
coefficient_derivatives
```

## Supported models and comparisons

Analytical confidence intervals are available for:

- first-order Markov models, where the next state depends only on the
  previous state and the covariates;
- full proportional-odds models with a reverse cumulative-logit link;
- frequentist models fitted with `orm_markov()` or `vglm_markov()`
  (including the `robcov_vglm` object that `vglm_markov()` returns);
- patient-level SOPs from `sops()` with conditional variance, and averaged
  SOPs from `avg_sops()` with conditional or unconditional variance;
- differences with `estimand = "sop"` or `estimand = "time_in_state"`; and
- numeric or factor visit times.

For other models, such as partial proportional-odds or second-order models,
or for `estimand = "time_benefit"`, use simulation or the bootstrap; see
`vignette("mvn-vs-bootstrap-orm-sops")`.
