---
title: "Bayesian SOPs with blrm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian SOPs with blrm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
RUN_BAYESIAN <- identical(
  Sys.getenv("MOSTR_RUN_BAYESIAN_VIGNETTE"),
  "true"
) && requireNamespace("rmsb", quietly = TRUE)
```

`blrm_markov()` fits a Bayesian transition model with `rmsb::blrm()`, which
samples from the posterior distribution of the coefficients using Stan. The
state occupancy probabilities (SOPs) and comparisons calculated from it carry
posterior uncertainty directly: mostr computes the SOPs for each posterior
draw of the coefficients and summarizes the results. No separate
`inferences()` step is needed.

This article fits a Bayesian model in which the effect of time may depart
from proportional odds, computes posterior SOPs, and checks the model
against the data.

Fitting the model takes a while, so the model-based code below runs only when
the environment variable `MOSTR_RUN_BAYESIAN_VIGNETTE` is `"true"` and the
rmsb package is installed. The example uses few MCMC iterations to keep
running time short; a real analysis needs more iterations and a check of
sampling diagnostics.

## Data

We simulate 200 patients over 14 days with `sim_actt2_markov()`, which uses
the eight-state NIAID scale described in `vignette("full-po-sops")`, and plot
the observed proportion of patients in each state.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)
library(rmsb)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260601
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"
SOP_DRAWS <- 200

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(raw_data, absorbing_state = 8)
```

```{r observed-sops}
plot_sops(raw_data)
```

## Fit the model

The formula includes the previous state and a treatment-by-time interaction,
with time modeled by a restricted cubic spline. Two further arguments relax
proportional odds for time:

- `ppo = ~ time` allows the effect of time to differ between cut-points of
  the ordinal scale;
- `cppo = function(y) as.numeric(y)` constrains that difference to change
  linearly with the state number, so it adds only one coefficient.

`cluster(id)` adds a random effect for each patient, which allows for
patient-level differences not explained by the covariates.

```{r fit-blrm, eval = RUN_BAYESIAN}
dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_blrm <- blrm_markov(
  y ~ yprev + rms::rcs(time, 4) * tx + cluster(id),
  ppo = ~ time,
  cppo = function(y) as.numeric(y),
  data = model_data,
  id_var = "id",
  iter = 300,
  warmup = 150,
  chains = 2,
  method = "sampling",
  refresh = 0,
  seed = SEED
)
fit_blrm
```

## Posterior SOPs

For Bayesian fits, `avg_sops()` computes the SOPs for each posterior draw and
then summarizes them. The arguments specific to Bayesian fits are:

- `n_draws = NULL` uses all stored posterior draws; a number uses a random
  subset of that size, selected reproducibly with `seed`;
- `include_re = TRUE` includes each patient's estimated random effect in the
  predictions;
- `posterior_summary = "mean"` reports the posterior mean. Unlike the median,
  the means across states always add up to 1;
- `conf_level = 0.90` requests 90% posterior intervals;
- `return_draws = TRUE` keeps the draw-level SOPs, which `get_draws()`
  extracts.

Averaging over patients for every posterior draw can need a lot of memory.
`options(mostr.blrm_avg_chunk_size = 100)` processes the draws in groups of
100 to limit it.

```{r avg-sops, eval = RUN_BAYESIAN}
options(mostr.blrm_avg_chunk_size = 100)

sop_blrm <- avg_sops(
  fit_blrm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_blrm$ylevels,
  absorb = ABSORB,
  include_re = TRUE,
  n_draws = NULL,
  seed = SEED,
  posterior_summary = "mean",
  conf_level = 0.90,
  return_draws = TRUE
)

head(sop_blrm)
```

```{r plot-sops, eval = RUN_BAYESIAN}
plot_sops(
  sop_blrm,
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "Posterior state occupancy probability"
  )
```

## Check the model against the data

As in `vignette("full-po-sops")`, each diagnostic is computed once from the
observed trajectories and once from the fitted model. For Bayesian fits,
the model-based summary is calculated separately for each posterior draw and
only then summarized across draws; `seed` makes the choice of draws reproducible. The transition
heatmaps use at most sixteen evenly spaced days.

```{r diagnostics-transitions-observed, eval = RUN_BAYESIAN, fig.width = 9, fig.height = 9}
diagnostic_times <- unique(round(seq(1, FOLLOW_UP, length.out = min(16L, FOLLOW_UP))))

plot_transitions(
  raw_data,
  times = diagnostic_times,
  y_levels = STATE_LEVELS,
  show_values = FALSE
) +
  labs(title = "Observed transition proportions")
```

```{r diagnostics-transitions-model, eval = RUN_BAYESIAN, fig.width = 9, fig.height = 9}
plot_transitions(
  fit_blrm,
  times = diagnostic_times,
  y_levels = fit_blrm$ylevels,
  absorb = ABSORB,
  seed = SEED,
  show_values = FALSE
) +
  labs(title = "Model-based transition proportions")
```

```{r diagnostics-correlation-observed, eval = RUN_BAYESIAN}
plot_correlation(
  fill_limits = c(0, 1),
  raw_data,
  times = seq_len(FOLLOW_UP),
  show_values = FALSE
) +
  labs(title = "Observed state correlations")
```

```{r diagnostics-correlation-model, eval = RUN_BAYESIAN}
plot_correlation(
  fill_limits = c(0, 1),
  fit_blrm,
  times = seq_len(FOLLOW_UP),
  y_levels = fit_blrm$ylevels,
  absorb = ABSORB,
  seed = SEED,
  show_values = FALSE
) +
  labs(title = "Model-based state correlations")
```

```{r diagnostics-variogram-observed, eval = RUN_BAYESIAN}
plot_variogram(
  raw_data,
  times = seq_len(FOLLOW_UP)
) +
  labs(title = "Observed correlation variogram")
```

```{r diagnostics-variogram-model, eval = RUN_BAYESIAN}
plot_variogram(
  fit_blrm,
  times = seq_len(FOLLOW_UP),
  y_levels = fit_blrm$ylevels,
  absorb = ABSORB,
  seed = SEED
) +
  labs(title = "Model-based correlation variogram")
```
