---
title: "Time as a Factor with orm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Time as a Factor with orm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

Many studies assess patients at a few scheduled visits rather than every day,
for example on days 3, 7, 14, and 21. The state between visits is then
unknown and cannot be reconstructed reliably. In this situation, a transition
model can describe the change from one *visit* to the next, with each visit
treated as a separate category of time.

This article shows how to

1. turn daily data into visit data and fit an `orm_markov()` model with time
   as a factor;
2. estimate state occupancy probabilities (SOPs) at each visit;
3. convert these to a calendar-day scale by interpolating between visits; and
4. compare the expected number of days spent in each state between arms.

## Visit data

We simulate daily states for 220 patients with `sim_actt2_markov()`, which
uses the eight-state NIAID scale described in `vignette("full-po-sops")`, and
keep only days 3, 7, 14, and 21. The observed SOPs of the full daily data are
shown for reference.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)

N_PATIENTS <- 220
SEED <- 20260531
VISIT_DAYS <- stats::setNames(c(3, 7, 14, 21), as.character(1:4))
STATE_LEVELS <- as.character(1:8)
ABSORB <- 8L

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  treatment_effect = log(0.8),
  seed = SEED
)

plot_sops(raw_data, facet_var = NULL, linetype_var = "tx")
```

`VISIT_DAYS` links each visit number to its calendar day. To build the visit
data, we

- keep the rows for the four visit days;
- recode `time` as a factor with levels `"1"` to `"4"` (the visit number),
  keeping the calendar day in `visit_day`;
- set `yprev` to the state at the previous *visit* rather than the previous
  day; and
- for the first visit, use the patient's state before day 1 as `yprev`, so
  that every patient's starting state is the same observed baseline state.

```{r visit-data}
baseline_state <- raw_data[raw_data$time == 1, c("id", "yprev")]
names(baseline_state)[2] <- "baseline_yprev"

visit_data <- raw_data[
  raw_data$time %in% unname(VISIT_DAYS),
  c("id", "tx", "time", "y", "yprev")
]
visit_data <- merge(
  visit_data,
  baseline_state,
  by = "id",
  all.x = TRUE,
  sort = FALSE
)
visit_data <- visit_data[order(visit_data$id, visit_data$time), ]
visit_data$visit_day <- visit_data$time
visit_data$time <- factor(
  match(visit_data$visit_day, unname(VISIT_DAYS)),
  levels = seq_along(VISIT_DAYS),
  labels = names(VISIT_DAYS)
)
visit_data$yprev <- ave(
  visit_data$y,
  visit_data$id,
  FUN = function(y) c(NA_integer_, y[-length(y)])
)
first_visit <- !duplicated(visit_data$id)
visit_data$yprev[first_visit] <- visit_data$baseline_yprev[first_visit]
visit_data$baseline_yprev <- NULL

model_data <- prepare_markov_data(
  visit_data,
  absorbing_state = ABSORB,
  factor_previous = TRUE
)

head(model_data)
```

## Fit the model and estimate SOPs at each visit

With time as a factor, the model estimates a separate effect for each visit,
so the intervals between visits may differ in length without any assumption
about how transition probabilities change from day to day.

mostr needs to know which visit is the first, because each patient's starting
state and covariates are taken from their row at that visit. For numeric time
this defaults to 1; for factor time, set it explicitly with
`first_followup_time = "1"`. Likewise, pass the visit labels, not calendar
days, to `times` in `avg_sops()`. The resulting SOPs refer to visits 1 to 4.

```{r fit-model}
dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_visit <- orm_markov(
  y ~ tx + time + yprev,
  data = model_data,
  id_var = "id",
  first_followup_time = "1",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

sop_visit <- avg_sops(
  fit_visit,
  variables = list(tx = c(0, 1)),
  times = names(VISIT_DAYS),
  y_levels = fit_visit$yunique,
  absorb = as.character(ABSORB)
) |>
  inferences(method = "delta", vcov = "unconditional")

head(sop_visit)
```

## Interpolate to calendar days

`interpolate_sops()` places the visit-scale SOPs on the calendar-day scale:

- `time_map = VISIT_DAYS` maps each visit label to its day;
- `baseline_time = 0` adds the observed distribution of starting states
  (`yprev` at the first visit) as the SOPs on day 0;
- `target_times = 1:21` requests every day from 1 to 21.

Between two known time points, each probability is interpolated linearly.
The day-0 baseline anchors the interpolation for days 1 and 2, before the
first visit, but is not itself part of the returned grid. Interpolated values
are a convenient summary, not model predictions for the unobserved days.

```{r plot-sops}
sop_days <- interpolate_sops(
  sop_visit,
  time_map = VISIT_DAYS,
  target_times = 1:max(VISIT_DAYS),
  baseline_time = 0
)

plot_sops(
  sop_days,
  time_var = "time",
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  scale_x_continuous(breaks = c(0, unname(VISIT_DAYS))) +
  labs(
    x = "Calendar day",
    y = "State occupancy probability"
  )
```

## Compare time in each state

`avg_comparisons()` accepts the same `time_map`, `target_times`, and
`baseline_time` arguments. With them, `estimand = "time_in_state"` measures
the expected number of calendar days, rather than visits, spent in each state
over days 1 to 21. `time_unit = "days"` labels the result. Each estimate is
treated minus control.

```{r}
cmp_days <- avg_comparisons(
  fit_visit,
  variables = list(tx = c(0, 1)),
  estimand = "time_in_state",
  state_sets = NULL,
  comparison = "difference",
  times = names(VISIT_DAYS),
  y_levels = fit_visit$yunique,
  absorb = as.character(ABSORB),
  time_map = VISIT_DAYS,
  target_times = 1:max(VISIT_DAYS),
  baseline_time = 0,
  time_unit = "days"
) |>
  inferences(method = "delta", vcov = "unconditional")

plot_comparisons(cmp_days) +
  labs(y = "Difference in time in state (days)")
```
