---
title: "Fractional Weighted Bootstrap for SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Fractional Weighted Bootstrap for SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

The ordinary patient bootstrap (`vignette("mvn-vs-bootstrap-orm-sops")`)
draws patients with replacement, so each resample leaves out about a third of
the patients and counts others several times. In small samples or subgroups,
a resample can then lack patients in a rare state, and the model cannot be
refitted, or a subgroup can be missing altogether.

The *fractional weighted bootstrap* (FWB) avoids this. Instead of resampling,
it keeps every patient in every replicate and gives each patient a random
positive weight, drawn from an exponential distribution and scaled to have
mean 1. For each replicate it

1. refits the transition model, applying each patient's weight to all of
   that patient's rows; and
2. averages the patient-level state occupancy probabilities (SOPs) using the
   same weights.

Like the ordinary bootstrap, the FWB reflects both uncertainty in the model
coefficients and which patients happened to be sampled.

This article uses the FWB to estimate SOPs within subgroups defined by the
patients' starting state, and to compare time benefit between arms in these
subgroups.

## Fit the model

We simulate 160 patients over 10 days with `sim_actt2_markov()`, which uses
the eight-state NIAID scale described in `vignette("full-po-sops")`. We record
each patient's starting state, the value of `yprev` on day 1, in a new column
`starting_state` for use as a subgroup.

```{r setup}
library(mostr)
library(ggplot2)
library(dplyr)
library(rms)

N_PATIENTS <- 160
FOLLOW_UP <- 10
N_BOOT <- 30
SEED <- 20260606
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = -0.03,
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
) |>
  group_by(id) |>
  mutate(starting_state = first(yprev)) |>
  ungroup()

fit <- orm_markov(
  y ~ tx + rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)
```

## Define SOPs by subgroup

`by = "starting_state"` asks `avg_sops()` to average the patient-level SOPs
separately within each starting state, after setting every patient to each
treatment value. Because the model was fitted with `orm_markov()`, the full
longitudinal data needed for refitting and each patient's starting row are
stored with the fit, so neither needs to be passed again.

```{r avg-sops}
sop_by_start <- avg_sops(
  fit,
  variables = "tx",
  by = "starting_state",
  times = seq_len(FOLLOW_UP),
  y_levels = fit$yunique,
  absorb = ABSORB
)

head(sop_by_start)
```

## Run the fractional weighted bootstrap

`method = "fwb"` selects the FWB. Within each replicate, the subgroup
averages use the same patient weights as the refit. We use only 30 replicates
to keep this example fast; a real analysis needs many more, typically 1,000
or more.

```{r run-bootstrap}
set.seed(SEED)
sop_fwb <- inferences(
  sop_by_start,
  method = "fwb",
  n_draws = N_BOOT,
  workers = 1,
  return_draws = TRUE
)
```

Each panel shows one starting state. Subgroups with few patients have wider
intervals.

```{r compare}
plot_sops(
  sop_fwb,
  facet_var = "starting_state",
  linetype_var = "tx"
)
```

## Compare time benefit by subgroup

Time benefit summarizes the whole ordinal scale in one number: on each day,
the probability that a patient would be in a better state under treatment
than under control, minus the probability of a worse state, summed over days.
See `vignette("many-levels-previous-state-spline")` for a fuller explanation.
With `by = "starting_state"`, it is reported separately for each starting
state.

```{r compare-time-benefit}
tb_fwb <- avg_comparisons(
  fit,
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "time_benefit",
  absorb = ABSORB,
  by = "starting_state"
) |>
  inferences(
    method = "fwb",
    n_draws = N_BOOT,
    workers = 1,
    return_draws = TRUE
  )

plot_comparisons(tb_fwb, color_var = "starting_state")
```
