---
title: "State Occupancy Probabilities with Proportional-Odds Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State Occupancy Probabilities with Proportional-Odds Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-options, include = FALSE}
source("_setup.R", local = TRUE)
```

Many clinical trials record a patient's condition repeatedly on an ordered
scale, for example once a day from "at home" to "dead". A natural way to
summarize such data is the *state occupancy probability* (SOP): the
probability that a patient is in a given state on a given day. Comparing SOPs
between treatment arms shows when, and in which states, a treatment makes a
difference.

mostr estimates SOPs from a *transition model*: an ordinal regression that
predicts tomorrow's state from today's state, treatment, time, and other
covariates. Chaining these one-day predictions together gives the probability
of each state on every later day. This article walks through the complete
workflow:

1. simulate and prepare longitudinal data;
2. fit a proportional-odds transition model;
3. estimate SOPs for each treatment arm, with confidence intervals;
4. check whether the fitted model reproduces features of the observed data.

We fit the same model twice, once with `vglm_markov()` (using the VGAM
package) and once with `orm_markov()` (using the rms package), to show that
both give the same answer.

## Example data

We simulate daily health states for 400 patients over 28 days. The simulation
model is based on data from the
[Adaptive COVID-19 Treatment Trial 2 (ACTT-2)](https://doi.org/10.1056/NEJMoa2031994),
which enrolled adults hospitalized with moderate to severe COVID-19,[^niaid]
so the simulated trajectories resemble those seen in the trial. We choose the
treatment effect ourselves: `treatment_effect = log(0.8)` means that, on any
day, treated patients have 0.8 times the odds of being in a worse state than
control patients who started the day in the same state.

States follow the trial's
[NIAID eight-point ordinal scale](https://www.fda.gov/media/144473/download#page=10).
Higher numbers indicate worse health:

| State | Meaning |
|---|---|
| 1 | Outside hospital, with usual activities unrestricted |
| 2 | Outside hospital, with restricted activities, oxygen at home, or both |
| 3 | In hospital, without oxygen and no further medical care needed |
| 4 | In hospital, without oxygen but still needing medical care |
| 5 | In hospital, needing supplemental oxygen |
| 6 | In hospital, needing high-flow oxygen or noninvasive ventilation |
| 7 | In hospital, needing invasive ventilation or extracorporeal membrane oxygenation (ECMO) |
| 8 | Death |

Death is an *absorbing* state: once a patient enters it, they stay there for
the rest of follow-up.

[^niaid]: We thank the National Institute of Allergy and Infectious Diseases
    (NIAID) for sharing the ACTT-2 data, which enabled us to create synthetic
    trajectories that mimic patterns observed in real patients. Researchers
    can apply for access through the
    [NIAID clinical-trial data portal](https://accessclinicaldata.niaid.nih.gov/study-viewer/clinical_trials/NCT04401579).

The simulated data have one row per patient and day, with the columns that
mostr expects by default: `id` (patient), `time` (day), `y` (state on that
day), `yprev` (state on the previous day), and `tx` (treatment, 0 or 1).
`prepare_markov_data()` turns them into model-ready data. It drops the rows
recorded after a patient has died, because those days carry no information
about transitions, and converts `yprev` to a factor so that each previous
state gets its own coefficient.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)
library(VGAM)

N_PATIENTS <- 400
FOLLOW_UP <- 28
SEED <- 20260527
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = log(0.8),
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
```

Before modeling, it helps to look at the data. `plot_sops()` applied to raw
trajectories shows the observed proportion of patients in each state on each
day, separately by arm.

```{r observed-sops}
plot_sops(raw_data, facet_var = NULL, linetype_var = "tx")
```

## Fit a transition model

Both models below predict the state on each day, `y`, from treatment, the
previous day's state, and a restricted cubic spline in time with four knots,
`rcs(time, 4)`, which lets transition probabilities change smoothly over
follow-up. The models assume *proportional odds*: each covariate shifts the
odds of being in a worse state by the same factor, whichever cut-point of the
scale we look at.

The formula syntax follows each backend. VGAM needs an ordered-factor
response and a `cumulative()` family, where `reverse = TRUE` models the odds of
worse (higher) states and `parallel = TRUE` requests proportional odds. The
rms model needs a `datadist` object, which rms uses to summarize predictors.
The `opt_method`, `scale`, and `maxit` arguments help `rms::orm()` converge
reliably.

In both wrappers, `id_var = "id"` tells mostr which rows belong to the same
patient. mostr uses it to compute standard errors that account for repeated
observations of the same patient, and it stores the patient data needed later
to estimate SOPs and their confidence intervals.

```{r fit-models}
fit_vglm <- vglm_markov(
  ordered(y) ~ tx + rcs(time, 4) + yprev,
  family = cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data,
  id_var = "id"
)

dd <- datadist(model_data)
options(datadist = "dd")

fit_orm <- orm_markov(
  y ~ tx + rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

summary(fit_vglm)
fit_orm
```

The two fits describe the same model, so their coefficients agree. The `tx`
coefficient is the log odds ratio for treatment; a negative value means that
treated patients tend to move to better states. Its estimate should be close
to the simulated value, $\log(0.8) \approx -0.22$.

## Estimate state occupancy probabilities

`avg_sops()` answers the question: *what would the probability of each state
be on each day if every patient in the trial had received control, and if
every patient had received treatment?* It does so in three steps:

1. For each patient, it sets `tx` to 0, starts from that patient's observed
   state before day 1, and applies the model's one-day transition
   probabilities day after day to obtain the patient's SOPs.
2. It repeats the calculation with `tx` set to 1.
3. It averages the patient-level SOPs within each arm.

Because both arms are averaged over the same patients, differences between
them reflect the treatment rather than differences in who was randomized to
each arm.

The main arguments are:

- `variables = list(tx = c(0, 1))`: the treatment values to compare;
- `times`: the days for which to report SOPs;
- `y_levels`: the states in the order used by the model;
- `absorb`: the absorbing state, so that patients who die stay dead.

`inferences()` then adds standard errors and 95% confidence intervals.
`method = "delta"` computes them analytically, which is fast and needs no
random draws. `vcov = "unconditional"` makes the intervals account both for
uncertainty in the model coefficients and for the particular patients who
happened to enroll. See `vignette("analytical-confidence-intervals")` for
details and alternatives.

```{r compute-sops}
sop_vglm <- avg_sops(
  fit_vglm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = factor(STATE_LEVELS, levels = STATE_LEVELS),
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

sop_orm <- avg_sops(
  fit_orm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_orm$yunique,
  absorb = ABSORB
) |>
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

combined_sops <- rbind(
  data.frame(backend = "VGAM vglm", as.data.frame(sop_vglm)),
  data.frame(backend = "rms orm", as.data.frame(sop_orm))
)
```

The result has one row per day, state, and treatment value, with the estimated
probability in `estimate` and its confidence limits in `conf.low` and
`conf.high`. Plotting both backends side by side confirms that they agree.

```{r plot-sops}
plot_sops(
  combined_sops,
  geom = "line",
  linetype_var = "tx",
  facet_var = c("backend")
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## How the probabilities are calculated

This section states the calculation behind `avg_sops()` precisely. You can
skip it on a first reading.

Let $i = 1,\ldots,n$ index patients and $t = 1,\ldots,d$ index follow-up days.
Patient $i$'s state on day $t$ is $Y_{it}$, one of $K$ ordered states
numbered $1,\ldots,K$, and $Y_{i0}$ is the observed state before the first
day. The vector $\mathbf{x}_{it}$ holds treatment, time, and any other
covariates on day $t$, and
$\mathbf{x}_{i,1:t} = (\mathbf{x}_{i1},\ldots,\mathbf{x}_{it})$ holds their
values up to day $t$. The SOP for state $l$ is

$$
S_{it}(l) =
  \Pr(Y_{it} = l \mid \mathbf{x}_{i,1:t}, Y_{i0}).
$$

The fitted model supplies the *transition probability*: the probability of
moving from state $k$ on day $t - 1$ to state $l$ on day $t$,

$$
T_{it}(l \mid k) =
  \Pr(Y_{it} = l \mid Y_{i,t-1} = k, \mathbf{x}_{it}).
$$

This is a *first-order Markov* model: given today's state, tomorrow's state
does not depend on earlier states. For an absorbing state $k$ we do not use
the model at all but set $T_{it}(l \mid k) = 1$ if $l = k$ and $0$ otherwise,
so a patient who has died stays dead.

To obtain the SOPs, add up the probability of arriving in state $l$ from each
possible previous state:

$$
S_{it}(l) = \sum_{k=1}^{K} S_{i,t-1}(k)\, T_{it}(l \mid k),
\qquad t = 1,\ldots,d,
$$

starting from $S_{i0}(l) = 1$ for the observed starting state $l = Y_{i0}$
and $0$ for every other state. In matrix form, if $\mathbf{s}_{it}$ is the row
vector of SOPs and $\mathbf{P}_{it}$ is the $K \times K$ matrix with
$T_{it}(l \mid k)$ in row $k$ and column $l$, each day's update is
$\mathbf{s}_{it} = \mathbf{s}_{i,t-1}\mathbf{P}_{it}$. mostr performs this
update for all patients at once and never has to predict transitions out of
an absorbing state; probability already in such a state is simply carried
forward.

Finally, write $S_{it}(l; z)$ for the SOP calculated with treatment set to
$z$ on every day. The arm-specific SOP reported by `avg_sops()` is the
average over patients,

$$
\bar{S}_t(l; z) =
  \frac{1}{n} \sum_{i=1}^n S_{it}(l; z).
$$

## Check the model against the data

SOPs are only as trustworthy as the transition model behind them. mostr's
diagnostic plots compute a summary once from the observed trajectories and
once from the fitted model, so that you can compare the two side by side.
Clear disagreements suggest that the model misses a feature of the data, for
example a time trend, an interaction, or dependence on states before the
previous day.

### Correlation between days

`plot_correlation()` shows the Spearman rank correlation between patients'
states on every pair of days. The first plot uses the observed data; the
second uses the correlations implied by the model. Both should show similar
patterns: high correlation between nearby days that weakens as days get
further apart. By default, each plot chooses its own color scale, so we set
`fill_limits = c(0, 1)` to give both plots the same scale. If any
correlation were negative, we would use `c(-1, 1)` instead, because values
outside the limits are shown in grey.

```{r correlation-diagnostics}
ec <- plot_correlation(
  fill_limits = c(0, 1),
  show_values = FALSE,
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y"
)

mc <- plot_correlation(
  fill_limits = c(0, 1),
  show_values = FALSE,
  fit_orm,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_orm$yunique,
  absorb = ABSORB
)

ec
mc
```

### Transitions between states

`plot_transitions()` draws, for selected days, a heatmap of the proportion of
patients who move from each previous state (rows) to each current state
(columns). Comparing the observed and model-based heatmaps shows whether the
model reproduces how patients move between states over follow-up. We show at
most sixteen evenly spaced days so each panel stays readable.

```{r transition-diagnostics, fig.width = 9, fig.height = 9}
diagnostic_times <- unique(round(seq(1, FOLLOW_UP, length.out = min(16L, FOLLOW_UP))))

et <- plot_transitions(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = diagnostic_times,
  show_values = FALSE
)

mt <- plot_transitions(
  fit_orm,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = diagnostic_times,
  y_levels = fit_orm$yunique,
  absorb = ABSORB,
  show_values = FALSE
)

et
mt
```

With `comparison = "difference"`, the same function shows how treatment
changes the model-based transition proportions: each tile is the proportion
under `tx = 1` minus the proportion under `tx = 0`.

```{r transition-difference, fig.width = 9, fig.height = 9}
td <- plot_transitions(
  fit_orm,
  comparison = "difference",
  variables = list(tx = c(0, 1)),
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = diagnostic_times,
  y_levels = fit_orm$yunique,
  absorb = ABSORB,
  show_values = FALSE
)
td
```

### How correlation decays with time

`plot_variogram()` condenses the correlation matrix into a single curve: the
correlation between two days plotted against the number of days between them.
If the observed correlations fall off more slowly than the model's, patients'
earlier history matters more than a first-order model allows, and a
second-order model (see `vignette("second-order-orm-sops")`) may fit better.

```{r variogram-diagnostics}
plot_variogram(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y"
) +
  coord_cartesian(ylim = c(0, 1))

plot_variogram(
  fit_orm,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_orm$yunique,
  absorb = ABSORB
) +
  coord_cartesian(ylim = c(0, 1))
```

## Where to go next

- `vignette("analytical-confidence-intervals")` explains the confidence
  intervals used here and how to compare arms directly with
  `avg_comparisons()`.
- `vignette("partial-po-vglm-sops")`, `vignette("second-order-orm-sops")`,
  `vignette("factor-time-orm-sops")`, and
  `vignette("many-levels-previous-state-spline")` show other model
  specifications.
- `vignette("mvn-vs-bootstrap-orm-sops")`,
  `vignette("fractional-weighted-bootstrap-sops")`, and
  `vignette("bayesian-rmsb-sops")` show simulation-based, resampling-based,
  and Bayesian uncertainty.
