---
title: "State Occupancy Probabilities with Proportional-Odds Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{State Occupancy Probabilities with Proportional-Odds Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r vignette-options, include = FALSE}
source("_setup.R", local = TRUE)
```

This vignette simulates daily health-state trajectories for **400 patients over
28 days**, using a model based on data from the
[Adaptive COVID-19 Treatment Trial 2 (ACTT-2)](https://doi.org/10.1056/NEJMoa2031994).
The trial enrolled adults hospitalized with moderate to severe COVID-19.[^niaid]
The simulated trajectories resemble patterns in the trial data, with the
treatment effect chosen for this example using `treatment_effect = log(0.8)`.

We estimate the probability of being in each health state on each day, called
a *state occupancy probability* (SOP). We fit a proportional-odds model in which
the next day's state depends on the current state, treatment, and time, using
`vglm_markov()` (which uses VGAM) and `orm_markov()` (which uses rms).

## Health-state scale

The states follow the trial's
[NIAID eight-point ordinal scale](https://www.fda.gov/media/144473/download#page=10).
Higher numbers indicate worse health:

| State | Meaning |
|---|---|
| 1 | Outside hospital, with usual activities unrestricted |
| 2 | Outside hospital, with restricted activities, oxygen at home, or both |
| 3 | In hospital, without oxygen and no further medical care needed |
| 4 | In hospital, without oxygen but still needing medical care |
| 5 | In hospital, needing supplemental oxygen |
| 6 | In hospital, needing high-flow oxygen or noninvasive ventilation |
| 7 | In hospital, needing invasive ventilation or extracorporeal membrane oxygenation (ECMO) |
| 8 | Death |

State 8 is absorbing: once a simulated patient reaches this state, they remain
there for the rest of follow-up.

[^niaid]: We thank the National Institute of Allergy and Infectious Diseases
    (NIAID) for sharing the ACTT-2 data, which enabled us to create synthetic
    trajectories that mimic patterns observed in real patients. Researchers
    can apply for access through the
    [NIAID clinical-trial data portal](https://accessclinicaldata.niaid.nih.gov/study-viewer/clinical_trials/NCT04401579).

## Data

```{r setup}
library(mostr)
library(ggplot2)
library(rms)
library(VGAM)

N_PATIENTS <- 400
FOLLOW_UP <- 28
SEED <- 20260527
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = log(0.8),
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
```

```{r observed-sops}
plot_sops(raw_data, facet_var = NULL, linetype_var = "tx")
```

## Recursion

Let $i = 1,\ldots,n$ index patients and $t = 1,\ldots,d$ index follow-up
visits. The ordinal outcome $Y_{it}$ takes one of $K$ states, numbered
$1,\ldots,K$; $Y_{i0}$ is the observed starting state. Let
$\mathbf{x}_{it}$ collect treatment, time, and other covariates, and write
$\mathbf{x}_{i,1:t} = (\mathbf{x}_{i1},\ldots,\mathbf{x}_{it})$ for their
values through visit $t$. The state occupancy probability (SOP) for state $l$ is

$$
S_{it}(l) =
  \Pr(Y_{it} = l \mid \mathbf{x}_{i,1:t}, Y_{i0}).
$$

For a first-order transition model, the conditional probability of moving from
state $k$ at time $t - 1$ to state $l$ at time $t$ is

$$
T_{it}(l \mid k) =
  \Pr(Y_{it} = l \mid Y_{i,t-1} = k, \mathbf{x}_{it}).
$$

Let $\mathcal{A}$ be the set of absorbing states. For $k \in \mathcal{A}$,
define $T_{it}(l \mid k) = \mathbb{1}\{l = k\}$, where the indicator
$\mathbb{1}\{\cdot\}$ is one when its condition holds and zero otherwise.
This keeps a patient in the same absorbing state at every later visit.

For a single patient, collect these probabilities in the $K \times K$ transition matrix
$\mathbf{P}_{it}$, where row $k$, column $l$ is
$P_{it,kl} = T_{it}(l \mid k)$. If
$\mathbf{s}_{it}$ is the row vector with entries $S_{it}(l)$, then the
matrix recursion is

$$
\mathbf{s}_{i0} = \mathbf{e}_{Y_{i0}}, \qquad
\mathbf{s}_{it} = \mathbf{s}_{i,t-1} \mathbf{P}_{it}, \quad t = 1,\ldots,d.
$$

Here $\mathbf{e}_{Y_{i0}}$ is a row vector with one in the starting state's
position and zero elsewhere.

The implementation batches this same calculation across patients. Write
$\mathbf{S}_t$ for the $n \times K$ matrix with entry $S_{it}(l)$ in row $i$,
column $l$. For each
non-absorbing previous state $k$, it predicts an $n \times K$ matrix
$\mathbf{P}^{(k)}_t$ whose $i,l$ entry is $T_{it}(l \mid k)$. Without
predicting transitions from absorbing states, the update is

$$
S_{it}(l) =
  \sum_{k \notin \mathcal{A}}
    S_{i,t-1}(k) T_{it}(l \mid k)
  + \mathbb{1}\{l \in \mathcal{A}\} S_{i,t-1}(l).
$$

Equivalently, let $\mathbf{S}_{t,l}$ and $\mathbf{P}^{(k)}_{t,l}$ denote
column $l$ of the corresponding matrices. In vector form over patients,

$$
\mathbf{S}_{t,l} =
  \sum_{k \notin \mathcal{A}}
    \mathbf{S}_{t-1,k} \odot \mathbf{P}^{(k)}_{t,l}
  + \mathbb{1}\{l \in \mathcal{A}\} \mathbf{S}_{t-1,l},
$$

where $\odot$ denotes elementwise multiplication. The last term carries forward
probability already in an absorbing state, adding it exactly once.

The marginal arm-specific SOP used below is the average of these patient-level
SOPs after setting treatment to the desired counterfactual arm $z$. Write
$S_{it}(l; z)$ for the SOP calculated with `tx` set to $z$ at every visit:

$$
\bar{S}_t(l; z) =
  n^{-1} \sum_{i=1}^n S_{it}(l; z).
$$

## Fit Models

```{r fit-models}
fit_vglm <- vglm_markov(
  ordered(y) ~ tx + rcs(time, 4) + yprev,
  family = cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data,
  id_var = "id"
)

dd <- datadist(model_data)
options(datadist = "dd")

fit_orm <- orm_markov(
  y ~ tx + rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

summary(fit_vglm)
fit_orm
```

## Compute And Plot SOPs

```{r compute-sops}
sop_vglm <- avg_sops(
  fit_vglm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = factor(STATE_LEVELS, levels = STATE_LEVELS),
  absorb = ABSORB
) |> 
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

sop_orm <- avg_sops(
  fit_orm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_orm$yunique,
  absorb = ABSORB
) |> 
  inferences(
    method = "delta", 
    vcov = "unconditional")

combined_sops <- rbind(
  data.frame(backend = "VGAM vglm", as.data.frame(sop_vglm)),
  data.frame(backend = "rms orm", as.data.frame(sop_orm))
)
```

```{r plot-sops}
plot_sops(
  combined_sops,
  geom = "line",
  linetype_var = "tx",
  facet_var = c("backend")
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## Diagnostics

```{r correlation-diagnostics}
ec <- plot_correlation(
  show_values = FALSE,
  raw_data, 
  id_var = "id",
  time_var = "time",
  y_var = "y")

mc <- plot_correlation(
  show_values = FALSE,
  fit_orm, 
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP), 
  y_levels = fit_orm$yunique,
  absorb = ABSORB)

ec
mc
```

We show up to sixteen evenly spaced visits so each transition heatmap remains readable.
The observed, fitted, and treatment-difference plots use the same visits.

```{r transition-diagnostics, fig.width = 9, fig.height = 9}
diagnostic_times <- unique(round(seq(1, FOLLOW_UP, length.out = min(16L, FOLLOW_UP))))

et <- plot_transitions(
  raw_data, 
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = diagnostic_times,
  show_values = FALSE)

mt <- plot_transitions(
  fit_orm, 
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = diagnostic_times,
  y_levels = fit_orm$yunique,
  absorb = ABSORB,
  show_values = FALSE)

et
mt

td <- plot_transitions(
  fit_orm, 
  comparison = "difference",
  variables = list(tx = c(0, 1)),
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = diagnostic_times,
  y_levels = fit_orm$yunique,
  absorb = ABSORB,
  show_values = FALSE)
td
```

```{r variogram-diagnostics}
plot_variogram(
  raw_data, 
  id_var = "id",
  time_var = "time",
  y_var = "y"
  ) +
  coord_cartesian(ylim = c(0, 1))

plot_variogram(
  fit_orm, 
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP), 
  y_levels = fit_orm$yunique,
  absorb = ABSORB) +
  coord_cartesian(ylim = c(0, 1))
```
