---
title: "Many Outcome Levels with a Previous-State Spline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Many Outcome Levels with a Previous-State Spline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

Some outcomes have many ordered levels, such as a quality-of-life score from
1 to 30. Treating the previous state as a factor then adds one coefficient
per level, 29 in this case, and the rarer levels are poorly estimated.
Instead, the previous state can enter the model as a number, with a
restricted cubic spline allowing its effect to be non-linear. This uses a
handful of coefficients however many levels there are.

This article fits such a model, checks it against the data, and compares
arms using three summaries: state occupancy probabilities (SOPs), expected
time in each state, and time benefit.

## Simulate an outcome with 30 levels

We simulate 220 patients over 14 days on a 30-level scale where higher is
worse and no state is absorbing. `sim_trajectories_markov()` generates each
day's state from an ordinal model whose linear predictor is supplied in
`lp_function`: it increases linearly with the previous state, decreases
slightly over time, and is shifted by `parameter = -0.25` for treated
patients. The 29 cut-points are spread so that, with a linear predictor of
zero, the 30 states are equally likely.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)

N_PATIENTS <- 220
FOLLOW_UP <- 14
N_QOL <- 30
STATE_LEVELS <- as.character(seq_len(N_QOL))
ABSORB <- NULL
SEED <- 20260530

set.seed(SEED)
baseline <- data.frame(
  id = seq_len(N_PATIENTS),
  yprev = sample(12:28, N_PATIENTS, replace = TRUE),
  tx = rbinom(N_PATIENTS, 1, 0.5)
)
raw_data <- sim_trajectories_markov(
  baseline_data = baseline,
  follow_up_time = FOLLOW_UP,
  intercepts = qlogis(seq_len(N_QOL - 1L) / N_QOL),
  absorbing_states = integer(),
  covariate_names = "tx",
  parameter = -0.25,
  lp_function = function(yprev, t, tx, parameter, extra_params) {
    0.08 * (as.numeric(as.character(yprev)) - 15) - 0.03 * t +
      parameter * tx
  },
  seed = SEED
)

plot_sops(raw_data, geom = "bar")
```

## Fit a spline in the previous state

`prepare_markov_data(..., factor_previous = FALSE)` keeps `yprev` numeric so
that it can enter the model through `rcs(yprev, 5)`, a restricted cubic spline
with five knots. Because no state is absorbing, `absorbing_state = NULL`.

```{r fit-model}
model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = NULL,
  factor_previous = FALSE
)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_many <- orm_markov(
  y ~ rms::rcs(time, 4) + tx + rms::rcs(yprev, 5),
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  penalty = 0,
  maxit = 100
)

fit_many
```

## Check the model against the data

As in `vignette("full-po-sops")`, we compare the correlations between days
in the observed data with those implied by the model, first as a matrix of
all pairs of days and then as a variogram against the gap between days.
Without an absorbing state, `absorb = NULL`. Some observed
correlations are slightly negative, so both correlation plots use
`fill_limits = c(-1, 1)`; a shared scale makes their colors comparable.

```{r diagnostics-correlation}
ec <- plot_correlation(
  fill_limits = c(-1, 1),
  show_values = FALSE,
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y"
)

mc <- plot_correlation(
  fill_limits = c(-1, 1),
  show_values = FALSE,
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB
)

ec
mc
```

```{r diagnostics-variogram}
plot_variogram(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y"
)

plot_variogram(
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB
)
```

## Estimate state occupancy probabilities

`p_var = "yprev"` names the previous-state column. It is the default, shown
here to make clear that mostr substitutes each possible previous state into
this numeric column, and hence into the spline, when chaining predictions.

```{r avg-sops}
sop_many <- avg_sops(
  fit_many,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  p_var = "yprev"
) |>
  inferences(method = "delta", vcov = "unconditional")
```

With 30 states, two views help. Stacked bars show the whole distribution on
each day, one panel per arm. Line plots, one panel per state, show how each
state's probability changes over time in each arm.

```{r plot-sops, fig.height = 8}
plot_sops(
  sop_many,
  geom = "bar",
  facet_var = "tx"
) +
  scale_x_continuous(breaks = c(1, 7, 14)) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )

plot_sops(
  sop_many,
  geom = "line",
  facet_var = "state",
  linetype_var = "tx"
) +
  scale_x_continuous(breaks = c(1, 7, 14)) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## Compare the arms

### Differences in state occupancy probabilities

`estimand = "sop"` gives, for each state and day, the SOP under treatment
minus the SOP under control.

```{r compare-sops, fig.height = 8}
sop_diff <- avg_comparisons(
  fit_many,
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "sop"
) |>
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

plot_comparisons(sop_diff, facet_var = "state_set")
```

### Differences in time in state

`estimand = "time_in_state"` sums the daily SOPs, giving the expected number
of days spent in each state over the 14 days. With many levels, grouping
states into bands is often easier to interpret. For example,
`state_sets = list(good = 1:10, average = 11:20, bad = 21:30)` would compare
time spent in three bands; here we keep all 30 states.

```{r compare-time-in-state}
tis_many <- avg_comparisons(
  fit_many,
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "time_in_state"
) |>
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

plot_comparisons(tis_many)
```

### Time benefit

Time benefit summarizes all states in a single number. On each day, imagine
drawing one state from each arm's predicted distribution for the same
patient. The daily benefit is the probability that the treated state is better
(lower) minus the probability that it is worse. Summing over days gives the
expected number of days benefit: the net number of days on which treatment
leaves the patient better off. Results are averaged over patients.

Analytical intervals are not yet available for time benefit, so we use
simulation from the coefficients' multivariate normal distribution. Only 50
draws are used here to keep the example fast; use more, such as 1,000, in a
real analysis.

```{r compare-time-benefit}
tb_many <- avg_comparisons(
  fit_many,
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "time_benefit"
) |>
  inferences(
    method = "mvn",
    n_draws = 50
  )

plot_comparisons(tb_many)
```
