---
title: "Many Outcome Levels with a Previous-State Spline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Many Outcome Levels with a Previous-State Spline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

When the outcome has many ordered levels, a categorical previous-state effect can become too large. We can instead model the previous state effect using a restricted cubic spline to allow for non-linear effects.

## Simulate Many Ordered States

```{r setup}
library(mostr)
library(ggplot2)
library(rms)

N_PATIENTS <- 220
FOLLOW_UP <- 14
N_QOL <- 30
STATE_LEVELS <- as.character(seq_len(N_QOL))
ABSORB <- NULL
SEED <- 20260530

set.seed(SEED)
baseline <- data.frame(
  id = seq_len(N_PATIENTS),
  yprev = sample(12:28, N_PATIENTS, replace = TRUE),
  tx = rbinom(N_PATIENTS, 1, 0.5)
)
raw_data <- sim_trajectories_markov(
  baseline_data = baseline,
  follow_up_time = FOLLOW_UP,
  intercepts = qlogis(seq_len(N_QOL - 1L) / N_QOL),
  absorbing_states = integer(),
  covariate_names = "tx",
  parameter = -0.25,
  lp_function = function(yprev, t, tx, parameter, extra_params) {
    0.08 * (as.numeric(as.character(yprev)) - 15) - 0.03 * t +
      parameter * tx
  },
  seed = SEED
)

plot_sops(raw_data, geom = "bar")
```

The generator conditions each transition on the previous observed state, time,
and treatment. It uses 29 ordered logistic thresholds and no absorbing state.

## Fit `rcs(yprev, 5)`

`prepare_markov_data(..., factor_previous = FALSE)` keeps `yprev` numeric so it
can enter the model as a smooth previous-state effect.

```{r fit-model}
model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = NULL,
  factor_previous = FALSE
)

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_many <- orm_markov(
  y ~ rms::rcs(time, 4) + tx + rms::rcs(yprev, 5),
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  penalty = 0,
  maxit = 100
)

fit_many
```

## Diagnostics

```{r}
ec <- plot_correlation(
  show_values = FALSE,
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y")

mc <- plot_correlation(
  show_values = FALSE,
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB)

ec
mc
```


```{r}
plot_variogram(
  raw_data,
  id_var = "id",
  time_var = "time",
  y_var = "y"
  )

plot_variogram(
  fit_many,
  id_var = "id",
  time_var = "time",
  y_var = "y",
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  absorb = ABSORB)
```

## SOPs

```{r avg-sops}
sop_many <- avg_sops(
  fit_many,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_many$yunique,
  p_var = "yprev"
) |> 
  inferences(method = "delta", vcov = "unconditional")
```

```{r plot-sops}
plot_sops(
  sop_many,
  geom = "bar",
  facet_var = "tx"
) +
  scale_x_continuous(breaks = c(1, 7, 14)) +
  labs(
    x = "Week",
    y = "State occupancy probability"
  )

plot_sops(
  sop_many,
  geom = "line",
  facet_var = "state",
  linetype_var = "tx"
) +
  scale_x_continuous(breaks = c(1, 7, 14)) +
  labs(
    x = "Week",
    y = "State occupancy probability"
  )
```

## Contrasts

We can show the difference in the SOPs from above over time.

```{r}
sc <- avg_comparisons(
  fit_many, 
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "sop"
) |> 
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

plot_comparisons(sc, facet_var = "state_set")
```

We can calculate contrasts for time in state or weeks benefit.

```{r}
c <- avg_comparisons(
  fit_many, 
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)), 
  estimand = "time_in_state"
  # state_sets = list(good = 1:10, average = 11:20, bad = 21:30)
) |> 
  inferences(
    method = "delta",
    vcov = "unconditional"
  )

plot_comparisons(c)
```

Analytical delta intervals do not yet support `time_benefit`, so one needs to use bootstrapping or coefficient draws here.

```{r}
tb <- avg_comparisons(
  fit_many, 
  times = seq_len(FOLLOW_UP),
  variables = list(tx = c(0, 1)),
  estimand = "time_benefit"
) |> 
  inferences(
    method = "mvn",
    n_draws = 50
  )

plot_comparisons(tb)
```
