---
title: "MVN Simulation Versus Bootstrap for orm_markov() SOPs"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MVN Simulation Versus Bootstrap for orm_markov() SOPs}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

Analytical confidence intervals (`vignette("analytical-confidence-intervals")`)
are not available for every model or summary. Two general alternatives work
with frequentist models fitted by `orm_markov()` or `vglm_markov()`:

1. **Multivariate-normal (MVN) simulation**, `method = "mvn"`, draws many
   coefficient vectors from a multivariate normal distribution centered on
   the estimates, with the model's covariance matrix. For each draw it
   recomputes the state occupancy probabilities (SOPs) over the same
   patients. The spread of the results reflects uncertainty in the
   coefficients only, with the patients' starting states and covariates
   treated as given, like the conditional variance of the analytical method.
2. **Patient bootstrap**, `method = "bootstrap"`, resamples patients with
   replacement, refits the model to each resample, and recomputes the
   averaged SOPs over the resampled patients. Its spread reflects both
   uncertainty in the coefficients and which patients happened to be sampled,
   like the unconditional variance of the analytical method.

MVN simulation is much faster, because it never refits the model. The
bootstrap relies less on the normal approximation to the coefficients and
also reflects variation in the patients averaged over, but refitting can take
a long time for large models.

This article applies both methods to the same fitted model and compares the
results. It uses only 50 draws so that it runs quickly; a real analysis needs
many more, typically 1,000 or more, for stable interval limits.

## Simulate data and fit the model

We simulate 200 patients over 10 days with `sim_actt2_markov()`, which uses
the eight-state NIAID scale described in `vignette("full-po-sops")`, and fit
a proportional-odds transition model.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)

N_PATIENTS <- 200
FOLLOW_UP <- 10
N_DRAWS <- 50
SEED <- 20260602
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = log(0.8),
  seed = SEED
)

model_data <- prepare_markov_data(raw_data, absorbing_state = as.integer(ABSORB))

dd <- rms::datadist(model_data)
options(datadist = "dd")

fit_orm <- orm_markov(
  y ~ tx + rms::rcs(time, 4) + yprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)
```

Both methods rely on what `orm_markov()` stores with the fit. With
`id_var = "id"`, it computes a covariance matrix that allows for correlation
between each patient's transitions, which MVN simulation draws from. It also
keeps the full longitudinal data, which the bootstrap resamples and refits,
and each patient's starting row, over which `avg_sops()` averages.

## Estimate SOPs

The point estimates are computed once, and both methods start from the same
object.

```{r avg-sops}
avg <- avg_sops(
  fit_orm,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_orm$yunique,
  absorb = ABSORB
)
```

## Run both methods

`workers` sets the number of parallel processes; more than one speeds up the
bootstrap considerably. `return_draws = TRUE` keeps the individual draws,
which `get_draws()` extracts for further analysis. Setting the seed makes the
draws reproducible.

```{r run-inference}
set.seed(SEED)
sop_mvn <- inferences(
  avg,
  method = "mvn",
  n_draws = N_DRAWS,
  workers = 1,
  return_draws = TRUE
)

set.seed(SEED)
sop_boot <- inferences(
  avg,
  method = "bootstrap",
  n_draws = N_DRAWS,
  workers = 1,
  return_draws = TRUE
)
```

## Compare the results

```{r compare}
comparison <- rbind(
  data.frame(method = "MVN simulation", as.data.frame(sop_mvn)),
  data.frame(method = "Patient bootstrap", as.data.frame(sop_boot))
)

plot_sops(
  comparison,
  geom = "line",
  facet_var = c("method"),
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

The point estimates are identical, because both come from the same fitted
model. The intervals differ because the two methods answer different
questions. MVN simulation measures how precisely the average is known for
these 200 patients. The bootstrap measures how precisely it is known for the
population they were sampled from, so it also captures variation in the
patients' starting states. The difference is largest in the first days,
when the SOPs depend most strongly on where patients started. Magirr et al.
(2025) discuss this distinction for covariate-adjusted treatment effects in
randomized trials.

## References

Magirr D, Wang C, Przybylski A, Baillie M. Estimating the variance of
covariate-adjusted estimators of average treatment effects in clinical trials
with binary endpoints. *Pharmaceutical Statistics*. 2025;24(4):e70021.
[doi:10.1002/pst.70021](https://doi.org/10.1002/pst.70021)
