---
title: "Partial Proportional-Odds SOPs with vglm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial Proportional-Odds SOPs with vglm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

A proportional-odds model assumes that each covariate shifts the odds of
being in a worse state by the same factor at every cut-point of the ordinal
scale. For example, the effect of time on the odds of being in state 2 or
worse is assumed to equal its effect on the odds of being in state 8 or
worse. When this assumption is doubtful for some covariates, a *partial*
proportional-odds model lets their effects differ between cut-points while
keeping proportional odds for the others.

This article fits both kinds of model to the same data with `vglm_markov()`:
a full proportional-odds model, and a partial proportional-odds model in which
the effect of time changes linearly across the cut-points. We compare how well
the two models fit, and how closely each reproduces the observed state
occupancy probabilities (SOPs), transitions, and correlations between days.

## Simulate data

We simulate 1,000 patients over 28 days with `sim_actt1_markov()`. Its
trajectories resemble those of the
[Adaptive COVID-19 Treatment Trial 1 (ACTT-1)](https://doi.org/10.1056/NEJMoa2007764),
which compared Remdesivir with placebo in adults hospitalized with COVID-19.
States follow the eight-point NIAID scale described in
`vignette("full-po-sops")`, from 1 (at home, no limitations) to 8 (death).
Treated patients (`tx = 1`) have 0.8 times the odds of being in a worse state
on any day.

The simulator itself is a partial proportional-odds model. For the cut-point
"state $k$ or worse", its linear predictor contains the term
$-0.0188 \times k \times t$ on day $t$. The coefficient is small, but it grows
with both the day and the cut-point: on day 28 it lowers the log odds of
death (cut-point $k = 8$) by $0.0188 \times 8 \times 28 \approx 4.2$, but those
of being in state 2 or worse by only about 1.1. We can therefore check which
model recovers this structure.

```{r setup}
library(mostr)
library(ggplot2)
library(patchwork)
library(rms)
library(VGAM)

N_PATIENTS <- 1000
FOLLOW_UP <- 28
SEED <- 20261009
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt1_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = log(0.8),
  seed = SEED
)

model_data <- prepare_markov_data(raw_data, absorbing_state = 8)
```

## Fit the proportional-odds model

Both models use treatment, the previous state, and a restricted cubic spline
in time with six knots, `rcs(time, 6)`, which places the knots at quantiles
of the observed days.

In VGAM, `cumulative(reverse = TRUE)` models the odds of worse (higher)
states, and `parallel = TRUE` requests proportional odds for every term.

```{r fit-po}
form <- ordered(y) ~ tx + rcs(time, 6) + yprev

fit_po <- vglm_markov(
  form,
  family = VGAM::cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data,
  id_var = "id"
)
```

## Specify which effects may vary

VGAM expresses partial proportional odds through *constraint matrices*. The
model has one linear predictor per cut-point; with eight states there are
seven, for "state 2 or worse" through "state 8 or worse". Each model term has
a constraint matrix with one row per cut-point and one column per coefficient
for that term:

- a single column of ones gives one coefficient shared by all cut-points,
  which is proportional odds;
- a second column with a score for each cut-point adds a coefficient that
  changes the term's effect in proportion to that score. With the scores
  $(0, 1, \ldots, 6)$, the effect of the term at cut-point "state $k$ or
  worse" is $\beta + (k - 2)\,\tau$. Here $\beta$ is the effect at the first
  cut-point, "state 2 or worse", and $\tau$ is the *linear deviation*: the
  change in effect from one cut-point to the next.

The easiest way to obtain a valid set of constraint matrices is to fit the
proportional-odds model first and edit its constraints. We change two things:

1. The spline in time is represented by several basis columns. We give the
   linear basis column the extra linear-deviation coefficient and keep the
   other basis columns proportional.
2. The fitted model lists a separate constraint for each dummy variable of the
   factor `yprev`. The constraints we supply must instead be named by the
   formula term, so we replace these entries with a single `yprev` entry with
   proportional odds.

```{r constraints}
constraints <- fit_po$vglm_fit@constraints

# The linear basis column of the time spline.
linear_time_term <- grep(
  "^rcs\\(time.*time$",
  names(constraints),
  value = TRUE
)[[1]]

# One row per cut-point: seven for eight states.
n_thresholds <- nrow(constraints[[linear_time_term]])

# Replace the per-dummy yprev constraints with one term-level entry.
yprev_terms <- grep("^yprev", names(constraints), value = TRUE)
constraints[yprev_terms] <- NULL
constraints[["yprev"]] <- cbind(PO_effect = rep(1, n_thresholds))

# Effect at the first cut-point plus a change of the same size per cut-point.
constraints[[linear_time_term]] <- cbind(
  PO_effect = rep(1, n_thresholds),
  linear_deviation = seq_len(n_thresholds) - 1
)
constraints[[linear_time_term]]
```

## Fit the partial proportional-odds model

We refit the model with these constraints. `parallel = FALSE` tells VGAM not
to impose proportional odds on every term, so that our constraint matrices
decide instead. Starting from the proportional-odds fit's linear predictors
(`etastart`) helps the fit converge. The final line stores the constraint
list itself, rather than the name `constraints`, in the model call, so that
code that later re-evaluates the call does not depend on this variable.

```{r fit-ppo}
fit_ppo <- vglm_markov(
  form,
  family = VGAM::cumulative(reverse = TRUE, parallel = FALSE),
  data = model_data,
  id_var = "id",
  constraints = constraints,
  etastart = fit_po$vglm_fit@predictors
)
fit_ppo$vglm_fit@call$constraints <- constraints
```

The two models share all coefficients except the linear deviation. VGAM
numbers the coefficients of a term by the columns of its constraint matrix,
so the linear time effect at the first cut-point ends in `:1` and the linear
deviation in `:2`. In the simulator, too, the time effect changes by
$-0.0188\,t$ from one cut-point to the next, so the deviation estimate
should be close to -0.0188. The treatment coefficient should be close to
$\log(0.8) \approx -0.22$:

```{r compare-coefficients}
coef_po <- coef(fit_po$vglm_fit)
coef_ppo <- coef(fit_ppo$vglm_fit)
deviation_name <- paste0(linear_time_term, ":2")

data.frame(
  coefficient = c("tx", "linear deviation of time"),
  PO = c(coef_po[["tx"]], NA),
  PPO = c(coef_ppo[["tx"]], coef_ppo[[deviation_name]]),
  simulated = c(log(0.8), -0.0188)
)
```

## Compare model fit

The Akaike information criterion, `AIC()`, measures how well a model fits
the data while penalizing each extra coefficient by 2 units; lower values are
better. The Bayesian information criterion, `BIC()`, uses a larger penalty
that grows with the number of observations. A difference of more than about
10 is usually regarded as strong evidence for the model with the lower value.
Because the proportional-odds model is the partial proportional-odds model
with the deviation set to zero, we can also compare them with a likelihood
ratio test, `VGAM::lrtest()`.

```{r model-fit}
data.frame(
  model = c("Proportional odds", "Partial proportional odds"),
  coefficients = c(length(coef_po), length(coef_ppo)),
  AIC = c(AIC(fit_po$vglm_fit), AIC(fit_ppo$vglm_fit)),
  BIC = c(BIC(fit_po$vglm_fit), BIC(fit_ppo$vglm_fit))
)

VGAM::lrtest(fit_ppo$vglm_fit, fit_po$vglm_fit)
```

Both criteria strongly favour the partial proportional-odds model.

## Compare state occupancy probabilities

`avg_sops()` estimates, for each arm, the probability of each state on each
day, averaged over the patients' observed starting states (see
`vignette("full-po-sops")`). We compare both models' estimates with the
observed proportion of patients in each state, separately for each arm. Only
point estimates are needed here, so we do not call `inferences()`.

```{r sops-models}
sop_args <- list(
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = STATE_LEVELS,
  absorb = ABSORB
)
sop_po <- do.call(avg_sops, c(list(fit_po), sop_args))
sop_ppo <- do.call(avg_sops, c(list(fit_ppo), sop_args))

observed <- as.data.frame(
  prop.table(
    table(
      tx = raw_data$tx,
      time = raw_data$time,
      state = factor(raw_data$y, levels = STATE_LEVELS)
    ),
    margin = c(1, 2)
  ),
  responseName = "estimate"
)

sop_comparison <- rbind(
  data.frame(source = "Observed", observed[, c("tx", "time", "state", "estimate")]),
  data.frame(source = "Proportional odds", as.data.frame(sop_po)[, c("tx", "time", "state", "estimate")]),
  data.frame(source = "Partial proportional odds", as.data.frame(sop_ppo)[, c("tx", "time", "state", "estimate")])
)
sop_comparison$time <- as.integer(as.character(sop_comparison$time))
sop_comparison$source <- factor(
  sop_comparison$source,
  levels = c("Observed", "Proportional odds", "Partial proportional odds")
)
```

```{r sops-plot, fig.width = 8, fig.height = 6}
ggplot(
  subset(sop_comparison, tx == 0),
  aes(time, estimate, colour = source, linetype = source)
) +
  geom_line() +
  facet_wrap(~state, scales = "free_y", labeller = label_both, ncol = 4) +
  labs(
    x = "Day",
    y = "State occupancy probability",
    colour = NULL,
    linetype = NULL,
    title = "Control arm"
  ) +
  theme(legend.position = "bottom")
```

The partial proportional-odds curves follow the observed proportions. The
proportional-odds curves deviate systematically: they underestimate deaths
during the first three weeks and misplace the occupancy of the hospital
states, most visibly state 4. The mean absolute difference from the observed
proportions over all days, states, and arms summarizes this:

```{r sops-error}
sop_wide <- reshape(
  sop_comparison,
  idvar = c("tx", "time", "state"),
  timevar = "source",
  direction = "wide"
)
c(
  `Proportional odds` = mean(abs(
    sop_wide$`estimate.Proportional odds` - sop_wide$estimate.Observed
  )),
  `Partial proportional odds` = mean(abs(
    sop_wide$`estimate.Partial proportional odds` - sop_wide$estimate.Observed
  ))
)
```



## Compare transitions

`plot_transitions()` shows, for selected days, the proportion of all
patients who move from each previous state (x-axis) to each current state
(y-axis). For the observed data these are counts divided by the number of
patients; for a model they are the expected proportions implied by its
transition probabilities.

```{r transitions-data}
transition_sources <- list(
  Observed = plot_transitions(
    raw_data,
    times = seq_len(FOLLOW_UP),
    y_levels = STATE_LEVELS
  ),
  `Proportional odds` = plot_transitions(
    fit_po,
    times = seq_len(FOLLOW_UP),
    y_levels = STATE_LEVELS,
    absorb = ABSORB
  ),
  `Partial proportional odds` = plot_transitions(
    fit_ppo,
    times = seq_len(FOLLOW_UP),
    y_levels = STATE_LEVELS,
    absorb = ABSORB
  )
)
transitions <- do.call(rbind, Map(
  function(plot, source) {
    data.frame(
      source = source,
      time = as.integer(as.character(plot$data$.time_key)),
      previous_state = plot$data$previous_state,
      state = plot$data$state,
      estimate = plot$data$estimate
    )
  },
  transition_sources,
  names(transition_sources)
))
transitions$source <- factor(transitions$source, levels = names(transition_sources))
```

```{r transitions-plot, fig.width = 9, fig.height = 7}
ggplot(
  subset(transitions, time %in% c(2, 7, 14, 28)),
  aes(previous_state, state, fill = estimate)
) +
  geom_tile(colour = "white") +
  facet_grid(source ~ time, labeller = labeller(time = function(x) paste("Day", x))) +
  scale_fill_viridis_c(trans = "sqrt", limits = c(0, NA)) +
  coord_equal() +
  labs(x = "Previous state", y = "Current state", fill = "Proportion")
```

Most patients stay in their previous state, so the diagonal dominates all
three rows. Differences between the models appear in the smaller
off-diagonal proportions. The plotted values are easier to compare directly
over time. For example, the proportion of patients who die on each day after
being ventilated the day before:

```{r transitions-death, fig.height = 3.5}
ggplot(
  subset(transitions, previous_state == "7" & state == "8"),
  aes(time, estimate, colour = source, linetype = source)
) +
  geom_line() +
  labs(
    x = "Day",
    y = "Proportion of patients",
    colour = NULL,
    linetype = NULL,
    title = "Moves from ventilation (state 7) to death (state 8)"
  )
```

The proportional-odds model underestimates these deaths in the first week
and overestimates them later.

## Compare correlations between days

`plot_correlation()` shows the Spearman rank correlation between patients'
states on every pair of days, from the observed data or implied by a model.

```{r correlation, fig.width = 10, fig.height = 4}
correlation_days <- seq(1, FOLLOW_UP, by = 3)

correlation_plots <- list(
  plot_correlation(
    raw_data,
    times = correlation_days,
    fill_limits = c(0, 1),
    show_values = FALSE
  ) +
    labs(title = "Observed"),
  plot_correlation(
    fit_po,
    times = correlation_days,
    y_levels = STATE_LEVELS,
    absorb = ABSORB,
    fill_limits = c(0, 1),
    show_values = FALSE
  ) +
    labs(title = "Proportional odds"),
  plot_correlation(
    fit_ppo,
    times = correlation_days,
    y_levels = STATE_LEVELS,
    absorb = ABSORB,
    fill_limits = c(0, 1),
    show_values = FALSE
  ) +
    labs(title = "Partial proportional odds")
)
wrap_plots(correlation_plots, nrow = 1) + plot_layout(guides = "collect")
```

All three plots look almost the same. Correlations between days depend
mainly on how strongly patients tend to stay in their previous state, which both models capture.

## Compare treatment effects on time in each state

`avg_comparisons()` with `estimand = "time_in_state"` sums each arm's daily
SOPs over follow-up. The result is the expected number of days spent in each
state, and the comparison reports treated minus control. A positive
difference for state 1 means that treated patients are expected to spend more
of the 28 days at home without limitations.

Analytical confidence intervals (`method = "delta"`) are available only for
full proportional-odds models, so for the partial proportional-odds model we
use `method = "mvn"`. It draws coefficient vectors from a multivariate normal
distribution centered on the estimates, with the model's covariance matrix,
recomputes the result for each draw, and takes percentiles. We use only 100
draws to keep this example fast; use more, such as 1,000, in a real analysis.

```{r time-in-state}
tis_args <- list(
  estimand = "time_in_state",
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = STATE_LEVELS,
  absorb = ABSORB
)

tis_po <- do.call(avg_comparisons, c(list(fit_po), tis_args)) |>
  inferences(method = "delta", vcov = "conditional")
tis_ppo <- do.call(avg_comparisons, c(list(fit_ppo), tis_args)) |>
  inferences(method = "mvn", n_draws = 100, seed = SEED)

tis_comparison <- rbind(
  data.frame(model = "Proportional odds", as.data.frame(tis_po)),
  data.frame(model = "Partial proportional odds", as.data.frame(tis_ppo))
)
tis_comparison$model <- factor(
  tis_comparison$model,
  levels = c("Proportional odds", "Partial proportional odds")
)
tis_comparison[, c("model", "state_set", "estimate", "conf.low", "conf.high")]
```

```{r time-in-state-plot, fig.height = 3.5}
ggplot(
  tis_comparison,
  aes(state_set, estimate, ymin = conf.low, ymax = conf.high, colour = model)
) +
  geom_hline(yintercept = 0, colour = "grey60") +
  geom_pointrange(position = position_dodge(width = 0.5)) +
  labs(
    x = "State",
    y = "Difference in days (treated - control)",
    colour = NULL
  ) +
  theme(legend.position = "bottom")
```

The two models give similar treatment effects even though their SOPs differ.
