---
title: "Partial Proportional-Odds SOPs with vglm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Partial Proportional-Odds SOPs with vglm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

A proportional-odds model assumes that each covariate shifts the odds of
being in a worse state by the same factor at every cut-point of the ordinal
scale. For example, the effect of time on the odds of being in state 2 or
worse is assumed to equal its effect on the odds of being in state 5 or
worse. When this assumption is doubtful for some covariates, a *partial*
proportional-odds model lets their effects differ between cut-points while
keeping proportional odds for the others.

This article fits such a model with `vglm_markov()`. We let the effect of time
change linearly across the cut-points and keep proportional odds for
treatment and the previous state. We then check the model, estimate state
occupancy probabilities (SOPs), and compare expected time in each state
between arms.

## Simulate data

We simulate 200 patients over 14 days with `sim_actt2_markov()`, which uses
the eight-state NIAID scale described in `vignette("full-po-sops")`. To keep
the extra coefficients well estimated in this small example, we combine the
two outpatient states (1 and 2) and the two hospital states without oxygen
(3 and 4). This leaves six states, with death as state 6.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)
library(VGAM)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260528
STATE_LEVELS <- as.character(1:6)
ABSORB <- "6"

# Map states 1-2 to 1, 3-4 to 2, and 5-8 to 3-6.
collapse_actt2_states <- function(data) {
  data$y <- ifelse(
    data$y %in% c(1L, 2L),
    1L,
    ifelse(data$y %in% c(3L, 4L), 2L, data$y - 2L)
  )
  data$yprev <- ifelse(
    data$yprev %in% c(1L, 2L),
    1L,
    ifelse(data$yprev %in% c(3L, 4L), 2L, data$yprev - 2L)
  )
  data
}

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = -0.03,
  seed = SEED
) |>
  collapse_actt2_states()

model_data <- prepare_markov_data(raw_data, absorbing_state = 6)
```

## Specify which effects may vary

VGAM expresses partial proportional odds through *constraint matrices*. The
model has one linear predictor per cut-point; with six states there are five.
Each model term has a constraint matrix with one row per cut-point and one
column per coefficient for that term:

- a single column of ones gives one coefficient shared by all cut-points,
  which is proportional odds;
- a second column $(0, 1, 2, 3, 4)$ adds a coefficient that changes the
  term's effect by the same amount from one cut-point to the next, a linear
  departure from proportional odds.

The easiest way to obtain a valid set of constraint matrices is to fit the
proportional-odds model first and edit its constraints. We change two things:

1. The spline in time, `rcs(time, 4)`, is represented by several basis
   columns. We give the linear basis column the extra "linear departure"
   coefficient and keep the other basis columns proportional.
2. The fitted model lists a separate constraint for each dummy variable of the
   factor `yprev`. The constraints we supply must instead be named by the
   formula term, so we replace these entries with a single `yprev` entry with
   proportional odds.

```{r fit-ppo}
form <- ordered(y) ~ tx + rcs(time, 4) + yprev

fit_po <- vglm_markov(
  form,
  family = VGAM::cumulative(reverse = TRUE, parallel = TRUE),
  data = model_data,
  id_var = "id"
)

constraints <- fit_po$vglm_fit@constraints

# The linear basis column of the time spline.
linear_time_term <- grep(
  "^rcs\\(time, 4\\).*time$",
  names(constraints),
  value = TRUE
)[[1]]

# One row per cut-point: five for six states.
n_thresholds <- nrow(constraints[[linear_time_term]])

# Replace the per-dummy yprev constraints with one term-level entry.
yprev_terms <- grep("^yprev", names(constraints), value = TRUE)
constraints[yprev_terms] <- NULL
constraints[["yprev"]] <- cbind(PO_effect = rep(1, n_thresholds))

# Shared effect plus a linear change across cut-points.
constraints[[linear_time_term]] <- cbind(
  PO_effect = rep(1, n_thresholds),
  linear_deviation = seq_len(n_thresholds) - 1
)
```

Printing the list shows the result: only the linear time term has two
columns.

```{r}
constraints
```

We refit the model with these constraints. `parallel = FALSE` tells VGAM not
to impose proportional odds on every term, so that our constraint matrices
decide instead. Starting from the proportional-odds fit's linear predictors
(`etastart`) helps the fit converge. The final line stores the constraint
list itself, rather than the name `constraints`, in the model call, so that
code that later re-evaluates the call does not depend on this variable.

```{r}
fit_ppo <- vglm_markov(
  form,
  family = VGAM::cumulative(reverse = TRUE, parallel = FALSE),
  data = model_data,
  id_var = "id",
  constraints = constraints,
  etastart = fit_po$vglm_fit@predictors
)
fit_ppo$vglm_fit@call$constraints <- constraints

summary(fit_ppo)
```

## Check the model against the data

As in `vignette("full-po-sops")`, each diagnostic is computed once from the
observed trajectories and once from the fitted model. Similar patterns
suggest the model captures how patients move between states; clear
differences point to a missing feature. The transition heatmaps use at most
sixteen evenly spaced days.

```{r diagnostics-transitions-observed, fig.width = 9, fig.height = 9}
diagnostic_times <- unique(round(seq(1, FOLLOW_UP, length.out = min(16L, FOLLOW_UP))))

plot_transitions(
  raw_data,
  times = diagnostic_times,
  y_levels = STATE_LEVELS,
  show_values = FALSE
) +
  labs(title = "Observed transition proportions")
```

```{r diagnostics-transitions-model, fig.width = 9, fig.height = 9}
plot_transitions(
  fit_ppo,
  times = diagnostic_times,
  y_levels = STATE_LEVELS,
  absorb = ABSORB,
  show_values = FALSE
) +
  labs(title = "Model-based transition proportions")
```

```{r diagnostics-correlation-observed}
plot_correlation(
  fill_limits = c(0, 1),
  raw_data,
  times = 1:FOLLOW_UP,
  show_values = FALSE
) +
  labs(title = "Observed state correlations")
```

```{r diagnostics-correlation-model}
plot_correlation(
  fill_limits = c(0, 1),
  fit_ppo,
  times = 1:FOLLOW_UP,
  y_levels = STATE_LEVELS,
  absorb = ABSORB,
  show_values = FALSE
) +
  labs(title = "Model-based state correlations")
```

```{r diagnostics-variogram-observed}
plot_variogram(
  raw_data,
  times = 1:FOLLOW_UP
) +
  labs(title = "Observed correlation variogram")
```

```{r diagnostics-variogram-model}
plot_variogram(
  fit_ppo,
  times = 1:FOLLOW_UP,
  y_levels = STATE_LEVELS,
  absorb = ABSORB
) +
  labs(title = "Model-based correlation variogram")
```

## Estimate state occupancy probabilities

Analytical confidence intervals (`method = "delta"`) are available only for
full proportional-odds models, so we use the default `method = "mvn"`. It
draws coefficient vectors from a multivariate normal distribution centered on
the estimates, with the model's covariance matrix, recomputes the SOPs for
each draw, and takes percentiles of the results. We use only 100 draws to
keep this example fast; use more, such as 1,000, in a real analysis.

```{r avg-sops}
sop_ppo <- avg_sops(
  fit_ppo,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = STATE_LEVELS,
  absorb = ABSORB
) |>
  inferences(n_draws = 100)

plot_sops(
  sop_ppo,
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## Compare time in each state

`avg_comparisons()` with `estimand = "time_in_state"` sums each arm's daily
SOPs over follow-up. The result is the expected number of days spent in each
state, and the comparison reports treated minus control. A positive
difference for state 1 means that treated patients are expected to spend more
of the 14 days in the best state.

```{r}
tis_ppo <- avg_comparisons(
  fit_ppo,
  estimand = "time_in_state",
  variables = list(tx = c(0, 1)),
  times = 1:FOLLOW_UP,
  y_levels = STATE_LEVELS,
  absorb = ABSORB
) |>
  inferences(n_draws = 100)

plot_comparisons(tis_ppo)
```
