---
title: "Second-Order SOPs with orm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Second-Order SOPs with orm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

A first-order transition model predicts today's state from yesterday's state
alone. Sometimes the state two days ago carries extra information: a patient
who has just improved may be more likely to keep improving than one who has
been stable. A *second-order* model captures this by including both previous
states as predictors.

This article fits a second-order model with `orm_markov()` and shows the one
extra argument needed downstream: `p2_var`, which names the column holding
the state two days back. Every function that computes model-based
probabilities (`avg_sops()`, `avg_comparisons()`, and the diagnostic plots)
needs it, so that it chains predictions through both previous states.

## Data

We simulate 200 patients over 14 days with `sim_actt2_markov()`, which uses
the eight-state NIAID scale described in `vignette("full-po-sops")`. The
simulated data contain the previous state, `yprev`, but not the state two
days back, so we construct that column, `ypprev`, by shifting `yprev` by one
day within each patient. The simulation records only one state before day 1,
so on day 1 we set `ypprev` equal to `yprev`.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)

N_PATIENTS <- 200
FOLLOW_UP <- 14
SEED <- 20260529
STATE_LEVELS <- as.character(1:8)
ABSORB <- "8"

raw_data <- sim_actt2_markov(
  n_patients = N_PATIENTS,
  follow_up_time = FOLLOW_UP,
  treatment_effect = log(0.8),
  seed = SEED
)

model_data <- prepare_markov_data(
  raw_data,
  absorbing_state = as.integer(ABSORB),
  factor_previous = TRUE
)
model_data <- model_data[order(model_data$id, model_data$time), ]

# State two days back: yprev shifted by one day, repeating the first value.
model_data$ypprev <- ave(
  model_data$yprev,
  model_data$id,
  FUN = function(x) c(x[1], head(x, -1))
)
model_data$ypprev <- factor(model_data$ypprev, levels = levels(model_data$yprev))

head(model_data[c("id", "time", "tx", "ypprev", "yprev", "y")])
```

## Fit the model

The model includes both previous states as factors. It also lets the effect
of each previous state change linearly over time through the `%ia%`
interaction terms, rms's operator for restricted interactions.

```{r fit-model}
dd <- datadist(model_data)
options(datadist = "dd")

fit_second <- orm_markov(
  y ~ tx + rcs(time, 4) + yprev + ypprev + time %ia% yprev + time %ia% ypprev,
  data = model_data,
  id_var = "id",
  opt_method = "LM",
  scale = TRUE,
  maxit = 100
)

fit_second
```

## Estimate state occupancy probabilities

Passing `p2_var = "ypprev"` switches `avg_sops()` from the first-order to the
second-order calculation described [below](#how-the-probabilities-are-calculated).
Analytical confidence intervals are available only for first-order models, so
we use simulation from the coefficients' multivariate normal distribution
(`method = "mvn"`, the default). We use only 50 draws to keep the example
fast; use more, such as 1,000, in a real analysis.

```{r avg-sops}
sop_second <- avg_sops(
  fit_second,
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_second$yunique,
  absorb = ABSORB,
  p2_var = "ypprev"
) |>
  inferences(n_draws = 50)

head(sop_second)
```

```{r plot-sops}
plot_sops(
  sop_second,
  geom = "line",
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "State occupancy probability"
  )
```

## Compare time in groups of states

`state_sets` groups states before comparing them. Here we compare the
expected number of days spent at home (states 1–2), in hospital without
high-flow oxygen or ventilation (3–5), on high-flow oxygen or ventilation
(6–7), and dead (8). Each estimate is treated minus control.

```{r}
tis_second <- avg_comparisons(
  fit_second,
  estimand = "time_in_state",
  state_sets = list(1:2, 3:5, 6:7, 8),
  variables = list(tx = c(0, 1)),
  times = seq_len(FOLLOW_UP),
  y_levels = fit_second$yunique,
  absorb = ABSORB,
  p2_var = "ypprev"
) |>
  inferences(n_draws = 100)

plot_comparisons(tis_second)
```

## How the probabilities are calculated

This section extends the calculation in `vignette("full-po-sops")` to two
previous states. You can skip it on a first reading.

Let $i = 1,\ldots,n$ index patients, $t = 1,\ldots,d$ index follow-up days,
and $Y_{it}$ be patient $i$'s state on day $t$, one of $K$ ordered states.
The vector $\mathbf{x}_{it}$ holds treatment, time, and other covariates on
day $t$, and $\mathbf{x}_{i,1:t}$ holds their values up to day $t$. Two
starting states are needed, $Y_{i,-1}$ and $Y_{i0}$; as noted above, this
example sets $Y_{i,-1} = Y_{i0}$. The state occupancy probability (SOP) for
state $l$ is

$$
S_{it}(l) =
  \Pr(Y_{it} = l \mid \mathbf{x}_{i,1:t}, Y_{i,-1}, Y_{i0}).
$$

The fitted model supplies the probability of moving to state $l$ given the
states $h$ two days back and $k$ one day back,

$$
T_{it}(l \mid h, k) =
  \Pr(Y_{it} = l \mid Y_{i,t-2} = h, Y_{i,t-1} = k, \mathbf{x}_{it}),
$$

obtained by setting `ypprev` to $h$, `yprev` to $k$, and `time` to $t$. If
$k$ is an absorbing state, the model is not used: $T_{it}(l \mid h, k)$ is 1
for $l = k$ and 0 otherwise, whatever $h$ is.

Because tomorrow depends on the last two states, the SOPs alone are not
enough to take the next step. Instead, the calculation tracks the joint
probability of the last two states,

$$
J_{it}(k,l) =
  \Pr(Y_{i,t-1} = k, Y_{it} = l \mid \mathbf{x}_{i,1:t}, Y_{i,-1}, Y_{i0}).
$$

On the first day, only the observed starting states are possible:

$$
J_{i1}(k,l) =
  \mathbb{1}\{k = Y_{i0}\}\, T_{i1}(l \mid Y_{i,-1}, Y_{i0}),
$$

where $\mathbb{1}\{\cdot\}$ is 1 when its condition holds and 0 otherwise. On
each later day, the new joint probability adds up the paths through every
possible state $h$ two days back:

$$
J_{it}(k,l) =
  \sum_{h=1}^{K}
    J_{i,t-1}(h,k)\, T_{it}(l \mid h,k),
\qquad t = 2,\ldots,d.
$$

The SOP is the probability of the current state regardless of the previous
one:

$$
S_{it}(l) =
  \sum_{k=1}^{K} J_{it}(k,l).
$$

As in the first-order case, mostr performs these updates for all patients at
once, carries forward probability in absorbing states without predicting from
them, and averages the patient-level SOPs within each treatment arm. It
computes the probabilities exactly rather than by simulating individual
trajectories.

## Check the model against the data

The diagnostics compare observed summaries with those implied by the model,
as explained in `vignette("full-po-sops")`. The model-based plots also need
`p2_var = "ypprev"`. Because a second-order model remembers more of each
patient's history, it can reproduce correlations that persist over longer
gaps than a first-order model allows; the variogram is the most direct place
to see this. The transition heatmaps use at most sixteen evenly spaced days.

```{r diagnostics-transitions-observed, fig.width = 9, fig.height = 9}
diagnostic_times <- unique(round(seq(1, FOLLOW_UP, length.out = min(16L, FOLLOW_UP))))

plot_transitions(
  raw_data,
  times = diagnostic_times,
  y_levels = STATE_LEVELS,
  show_values = FALSE
) +
  labs(title = "Observed transition proportions")
```

```{r diagnostics-transitions-model, fig.width = 9, fig.height = 9}
plot_transitions(
  fit_second,
  times = diagnostic_times,
  y_levels = fit_second$yunique,
  absorb = ABSORB,
  p2_var = "ypprev",
  show_values = FALSE
) +
  labs(title = "Model-based transition proportions")
```

```{r diagnostics-correlation-observed}
plot_correlation(
  fill_limits = c(0, 1),
  raw_data,
  times = seq_len(FOLLOW_UP),
  show_values = FALSE
) +
  labs(title = "Observed state correlations")
```

```{r diagnostics-correlation-model}
plot_correlation(
  fill_limits = c(0, 1),
  fit_second,
  times = seq_len(FOLLOW_UP),
  y_levels = fit_second$yunique,
  absorb = ABSORB,
  p2_var = "ypprev",
  show_values = FALSE
) +
  labs(title = "Model-based state correlations")
```

```{r diagnostics-variogram-observed}
plot_variogram(
  raw_data,
  times = seq_len(FOLLOW_UP)
) +
  labs(title = "Observed correlation variogram")
```

```{r diagnostics-variogram-model}
plot_variogram(
  fit_second,
  times = seq_len(FOLLOW_UP),
  y_levels = fit_second$yunique,
  absorb = ABSORB,
  p2_var = "ypprev"
) +
  labs(title = "Model-based correlation variogram")
```
