---
title: "Waning Treatment Effects with orm_markov()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Waning Treatment Effects with orm_markov()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
source("_setup.R", local = TRUE)
```

In the other vignettes we have assumed that the treatment has the same effect on
every day of follow-up, which usually not true. What happens to the estimated state occupancy
probabilities (SOPs) and time-in-state comparisons if, in truth, the effect
wears off?

We simulate daily health states for 400 patients over 28 days with a model
based on data from the
[Adaptive COVID-19 Treatment Trial 2 (ACTT-2)](https://doi.org/10.1056/NEJMoa2031994),
which enrolled adults hospitalized with moderate to severe COVID-19. States
follow the trial's
[NIAID eight-point ordinal scale](https://www.fda.gov/media/144473/download#page=10),
from unrestricted activities outside hospital (1) to death (8), and death is
absorbing. In all four scenarios, treated patients start with an odds ratio of
0.8 for moving to a worse state:

- **constant**: the odds ratio stays at 0.8 through day 28;
- **plateau 15**, **plateau 10**, and **plateau 5**: the odds ratio stays at
  0.8 through day 15, 10, or 5, and then returns toward 1, with the log odds
  ratio halving every 3 days.

To each simulated dataset we fit two `orm_markov()` transition models:

- **constant tx**, `y ~ tx + rcs(time, 4) + yprev`, which assumes a
  treatment effect that does not change over time; and
- **tx-by-time**, `y ~ tx * rcs(time, 3) + yprev`, which lets the treatment
  effect change smoothly over time.

For both, confidence intervals and p-values come from
`inferences(method = "delta", vcov = "unconditional")`. This analytical method
accounts for uncertainty in the coefficients and for which patients were
sampled; see `vignette("analytical-confidence-intervals")`. Unconditional
variance is the default for averaged results, but we request it explicitly.

## Time course of the simulated effect

Let $t = 1,\ldots,28$ index follow-up days and $D$ be the last day of the
plateau. The simulated log odds ratio on day $t$ is

$$
\beta(t) = \log(0.8) \times
  \begin{cases}
    1, & t \le D, \\
    \exp(-k(t - D)), & t > D,
  \end{cases}
  \qquad k = \log(2) / 3.
$$

where $k = \log(2)/3$ makes the log odds ratio halve every 3 days. The
constant scenario uses $D = 28$; the waning scenarios use
$D \in \{15, 10, 5\}$.

`sim_actt2_markov()` can only let an effect decay from day 1 onward (through
`treatment_effect_decay`), so the code below builds the simulation from the
general generator `sim_trajectories_markov()`. The function `plateau_lp()`
computes the linear predictor from the ACTT-2 coefficients for time and
previous state, plus the treatment effect with the plateau and decay above.
The plot shows the resulting odds ratio over time in each scenario.

```{r setup}
library(mostr)
library(ggplot2)
library(rms)

N_PATIENTS <- 400
FOLLOW_UP <- 28
KEY_DAYS <- c(7L, 14L, 21L, 28L)
N_REP <- 10
SEED <- 20260910
HALF_LIFE <- 3
LOG_OR <- log(0.8)
ABSORB <- "8"

plateau_lp <- function(yprev, t, tx, parameter, extra_params) {
  plateau <- extra_params[["plateau"]]
  rate <- log(2) / extra_params[["half_life"]]
  taper <- ifelse(t <= plateau, 1, exp(-rate * (t - plateau)))
  actt2 <- mostr:::actt2_markov_parameters()
  basis <- mostr:::actt_markov_rcs_basis(t)
  day_names <- c("day", "day'", "day''", "day'''", "day''''")
  time_effect <- drop(basis %*% actt2$extra_params[day_names])
  previous_state_effect <- actt2$extra_params[
    paste0("yprev=", as.character(yprev))
  ]
  as.numeric(time_effect + previous_state_effect + parameter * taper * tx)
}

sim_plateau <- function(n_patients, plateau, seed) {
  set.seed(seed)
  actt2 <- mostr:::actt2_markov_parameters()
  baseline_data <- data.frame(
    id = seq_len(n_patients),
    yprev = sample(
      actt2$baseline_states,
      size = n_patients,
      replace = TRUE,
      prob = actt2$baseline_probabilities
    ),
    tx = stats::rbinom(n_patients, size = 1L, prob = 0.5)
  )
  sim_trajectories_markov(
    baseline_data = baseline_data,
    follow_up_time = FOLLOW_UP,
    intercepts = rev(actt2$orm_intercepts),
    lp_function = plateau_lp,
    extra_params = c(
      actt2$extra_params,
      plateau = plateau,
      half_life = HALF_LIFE
    ),
    parameter = LOG_OR,
    absorbing_states = 8L,
    covariate_names = "tx"
  )
}

scenarios <- data.frame(
  scenario = c("constant", "plateau 15", "plateau 10", "plateau 5"),
  plateau = c(FOLLOW_UP, 15, 10, 5)
)

effect_curve <- expand.grid(
  day = seq_len(FOLLOW_UP),
  plateau = scenarios$plateau
)
rate <- log(2) / HALF_LIFE
effect_curve$odds_ratio <- exp(LOG_OR * ifelse(
  effect_curve$day <= effect_curve$plateau,
  1,
  exp(-rate * (effect_curve$day - effect_curve$plateau))
))
effect_curve$scenario <- scenarios$scenario[
  match(effect_curve$plateau, scenarios$plateau)
]
effect_curve$scenario <- factor(
  effect_curve$scenario,
  levels = scenarios$scenario
)

ggplot(effect_curve, aes(x = day, y = odds_ratio, color = scenario)) +
  geom_line(linewidth = 0.7) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  scale_color_viridis_d() +
  labs(
    x = "Day",
    y = "Simulated odds ratio (treated vs control)",
    color = "Scenario"
  )
```

With a 3-day half-life, the effect has almost vanished by day 28 in every
waning scenario; even after a 15-day plateau, the odds ratio on day 28 is
about 0.99.

## True SOPs and time in state

To measure bias we need the true values. Because we know the simulation
model, we can compute them exactly rather than by simulation. Starting from
the distribution of states before day 1, each day's SOPs are the previous
day's SOPs multiplied by that day's transition matrix, as described in
`vignette("full-po-sops")`. The true difference in expected time in a state
is the sum of the daily SOP differences between arms.

```{r truth}
true_sops <- function(plateau) {
  actt2 <- mostr:::actt2_markov_parameters()
  intercepts <- rev(actt2$orm_intercepts)
  day_coefs <- actt2$extra_params[
    c("day", "day'", "day''", "day'''", "day''''")
  ]
  previous_coefs <- unname(
    actt2$extra_params[paste0("yprev=", 1:7)]
  )
  rate <- log(2) / HALF_LIFE
  log_or_day <- function(t) {
    if (t <= plateau) LOG_OR else LOG_OR * exp(-rate * (t - plateau))
  }
  start <- numeric(8)
  start[actt2$baseline_states] <- actt2$baseline_probabilities
  arms <- list()
  for (z in c(0L, 1L)) {
    occupancy <- start
    daily <- matrix(NA_real_, nrow = FOLLOW_UP, ncol = 8)
    for (t in seq_len(FOLLOW_UP)) {
      basis <- mostr:::actt_markov_rcs_basis(t)
      time_effect <- drop(basis %*% day_coefs)
      transitions <- matrix(0, nrow = 8, ncol = 8)
      for (previous in 1:7) {
        linear_predictor <- time_effect +
          previous_coefs[previous] +
          log_or_day(t) * z
        cumulative <- stats::plogis(intercepts + linear_predictor)
        state_probs <- c(cumulative, 1) - c(0, cumulative)
        transitions[previous, ] <- rev(state_probs)
      }
      transitions[8, 8] <- 1
      occupancy <- drop(occupancy %*% transitions)
      daily[t, ] <- occupancy
    }
    arms[[as.character(z)]] <- daily
  }
  arms
}

truth <- lapply(scenarios$plateau, true_sops)
names(truth) <- scenarios$scenario

true_tis <- do.call(rbind, lapply(scenarios$scenario, function(scenario) {
  arms <- truth[[scenario]]
  data.frame(
    scenario = scenario,
    state = factor(as.character(1:8), levels = as.character(1:8)),
    truth = colSums(arms[["1"]]) - colSums(arms[["0"]])
  )
}))
true_tis$scenario <- factor(true_tis$scenario, levels = scenarios$scenario)

true_sop_day28 <- do.call(rbind, lapply(scenarios$scenario, function(scenario) {
  arms <- truth[[scenario]]
  data.frame(
    scenario = scenario,
    state = factor(as.character(1:8), levels = as.character(1:8)),
    truth = arms[["1"]][FOLLOW_UP, ] - arms[["0"]][FOLLOW_UP, ]
  )
}))
true_sop_day28$scenario <- factor(
  true_sop_day28$scenario,
  levels = scenarios$scenario
)

true_long <- do.call(rbind, lapply(scenarios$scenario, function(scenario) {
  arms <- truth[[scenario]]
  do.call(rbind, lapply(c("0", "1"), function(z) {
    daily <- arms[[z]]
    data.frame(
      day = rep(seq_len(FOLLOW_UP), times = 8),
      state = factor(
        rep(as.character(1:8), each = FOLLOW_UP),
        levels = as.character(1:8)
      ),
      estimate = as.vector(daily),
      scenario = scenario,
      arm = ifelse(z == "1", "treated", "control")
    )
  }))
}))
true_long$scenario <- factor(true_long$scenario, levels = scenarios$scenario)

plot_sops(
  true_long,
  time_var = "day",
  facet_var = "scenario",
  linetype_var = "arm"
) +
  labs(
    x = "Day",
    y = "State occupancy probability",
    title = "True SOPs by arm and scenario"
  )

true_tis
```

The table lists the true difference in expected days spent in each state,
treated minus control. These differences shrink as the plateau shortens:
the less time treatment works at full strength, the less the arms diverge.

## One illustrative dataset per scenario

Before running the full simulation, we analyze one dataset per scenario to
show the workflow. Each model is fitted with `orm_markov(id_var = "id")`, so
that the unconditional variance can account for the sampling of patients.
`avg_comparisons()` with `estimand = "sop"` gives the daily SOP differences
between arms, treated minus control.

```{r illustrate}
illustrate <- function(plateau, seed) {
  raw_data <- sim_plateau(N_PATIENTS, plateau, seed)
  model_data <- prepare_markov_data(
    raw_data,
    absorbing_state = as.integer(ABSORB)
  )
  dd <<- rms::datadist(model_data)
  options(datadist = "dd")
  fit_constant <- orm_markov(
    y ~ tx + rcs(time, 4) + yprev,
    data = model_data,
    id_var = "id",
    opt_method = "LM",
    scale = TRUE,
    maxit = 100
  )
  fit_interaction <- orm_markov(
    y ~ tx * rcs(time, 3) + yprev,
    data = model_data,
    id_var = "id",
    opt_method = "LM",
    scale = TRUE,
    maxit = 100
  )
  sop_one <- function(fit) {
    avg_comparisons(
      fit,
      variables = list(tx = c(0, 1)),
      estimand = "sop",
      state_sets = NULL,
      comparison = "difference",
      times = 1:FOLLOW_UP,
      y_levels = fit$yunique,
      absorb = ABSORB
    ) |>
      inferences(method = "delta", vcov = "unconditional")
  }
  list(
    raw_data = raw_data,
    constant = sop_one(fit_constant),
    interaction = sop_one(fit_interaction)
  )
}

illustrated <- Map(
  illustrate,
  scenarios$plateau,
  c(101L, 102L, 103L, 104L)
)
names(illustrated) <- scenarios$scenario

plot_sops(
  illustrated[["constant"]]$raw_data,
  facet_var = NULL,
  linetype_var = "tx"
) +
  labs(
    x = "Day",
    y = "State occupancy probability",
    title = "Observed SOPs, constant scenario (one dataset)"
  )

sop_curves <- do.call(rbind, lapply(scenarios$scenario, function(scenario) {
  rbind(
    data.frame(
      scenario = scenario,
      model = "constant tx",
      as.data.frame(illustrated[[scenario]]$constant)
    ),
    data.frame(
      scenario = scenario,
      model = "tx-by-time",
      as.data.frame(illustrated[[scenario]]$interaction)
    )
  )
}))
sop_curves$scenario <- factor(sop_curves$scenario, levels = scenarios$scenario)
sop_curves$model <- factor(
  sop_curves$model,
  levels = c("constant tx", "tx-by-time")
)
sop_home <- sop_curves[sop_curves$state_set == "1", ]

ggplot(
  sop_home,
  aes(x = time, y = estimate, color = model, fill = model)
) +
  geom_ribbon(
    aes(ymin = conf.low, ymax = conf.high),
    alpha = 0.15,
    color = NA
  ) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~scenario) +
  scale_color_viridis_d() +
  scale_fill_viridis_d() +
  labs(
    x = "Day",
    y = "SOP difference, state 1 (treated minus control)",
    color = "Fitted model",
    fill = "Fitted model"
  )
```

The plot shows the difference in the probability of being in state 1
(outside hospital, without restrictions) for each fitted model. The
constant-tx model assumes the same treatment effect on every day, whereas the
tx-by-time model can let the effect fade where the simulated effect fades.
The next section measures the consequences systematically.

## Simulation: power and bias

We simulate `r N_REP` datasets per scenario, each with `r N_PATIENTS`
patients. For each dataset we fit both models and compute:

- SOP differences, treated minus control, on days 7, 14, 21, and 28; and
- differences in expected time in each state over days 1 to 28.

Although we report SOP differences on only four days, they must be requested
for every day. `avg_sops()` and `avg_comparisons()` apply the transition
model one step per requested time, so requesting only days 7, 14, 21, and 28
would treat day 7 as the first day after baseline, and so on. We therefore
compute all 28 days and keep the four of interest. The expected time in each
state likewise sums all 28 daily SOPs.

We summarize each estimate by two measures:

- **power**: the proportion of datasets in which the two-sided p-value for
  no difference is below 0.05. `inferences(..., null = 0)` adds Wald tests
  of a zero difference;
- **bias**: the average estimate minus the true value.

With only $R$ = `r N_REP` datasets, both measures are themselves uncertain.
We report their Monte Carlo standard errors (MCSE): for power $p$, this is
$\sqrt{p(1 - p)/R}$; for bias, it is the standard deviation of the estimates
divided by $\sqrt{R}$.

The simulation takes several minutes, so the chunk is cached.

```{r simulate, cache = TRUE}
FULL_DAYS <- seq_len(FOLLOW_UP)
truth_key_days <- do.call(rbind, lapply(scenarios$scenario, function(scenario) {
  arms <- truth[[scenario]]
  data.frame(
    scenario = scenario,
    state = factor(as.character(1:8), levels = as.character(1:8)),
    truth = colSums(arms[["1"]][FULL_DAYS, , drop = FALSE]) -
      colSums(arms[["0"]][FULL_DAYS, , drop = FALSE])
  )
}))
truth_key_days$scenario <- factor(
  truth_key_days$scenario,
  levels = scenarios$scenario
)

truth_sop_key <- do.call(rbind, lapply(scenarios$scenario, function(scenario) {
  arms <- truth[[scenario]]
  pieces <- lapply(KEY_DAYS, function(day) {
    data.frame(
      scenario = scenario,
      time = day,
      state = factor(as.character(1:8), levels = as.character(1:8)),
      truth = arms[["1"]][day, ] - arms[["0"]][day, ]
    )
  })
  do.call(rbind, pieces)
}))
truth_sop_key$scenario <- factor(
  truth_sop_key$scenario,
  levels = scenarios$scenario
)

analyze_replicate <- function(plateau, seed) {
  raw_data <- sim_plateau(N_PATIENTS, plateau, seed)
  model_data <- prepare_markov_data(
    raw_data,
    absorbing_state = as.integer(ABSORB)
  )
  dd <<- rms::datadist(model_data)
  options(datadist = "dd")
  fits <- list(
    "constant tx" = orm_markov(
      y ~ tx + rcs(time, 4) + yprev,
      data = model_data,
      id_var = "id",
      opt_method = "LM",
      scale = TRUE,
      maxit = 100
    ),
    "tx-by-time" = orm_markov(
      y ~ tx * rcs(time, 3) + yprev,
      data = model_data,
      id_var = "id",
      opt_method = "LM",
      scale = TRUE,
      maxit = 100
    )
  )
  out <- list()
  for (model in names(fits)) {
    fit <- fits[[model]]
    sop <- avg_comparisons(
      fit,
      variables = list(tx = c(0, 1)),
      estimand = "sop",
      state_sets = NULL,
      comparison = "difference",
      times = FULL_DAYS,
      y_levels = fit$yunique,
      absorb = ABSORB
    ) |>
      inferences(method = "delta", vcov = "unconditional", null = 0)
    tis <- avg_comparisons(
      fit,
      variables = list(tx = c(0, 1)),
      estimand = "time_in_state",
      state_sets = NULL,
      comparison = "difference",
      times = FULL_DAYS,
      y_levels = fit$yunique,
      absorb = ABSORB
    ) |>
      inferences(method = "delta", vcov = "unconditional", null = 0)
    sop_df <- as.data.frame(sop)
    sop_df <- sop_df[sop_df$time %in% KEY_DAYS, ]
    sop_df$estimand <- "sop"
    sop_df$key <- paste(sop_df$time, sop_df$state_set, sep = "|")
    tis_df <- as.data.frame(tis)
    tis_df$estimand <- "time_in_state"
    tis_df$time <- NA_integer_
    tis_df$key <- paste("tis", tis_df$state_set, sep = "|")
    out[[model]] <- rbind(
      data.frame(
        model = model,
        estimand = sop_df$estimand,
        time = sop_df$time,
        state = sop_df$state_set,
        key = sop_df$key,
        estimate = sop_df$estimate,
        std.error = sop_df$std.error,
        p.value = sop_df$p.value
      ),
      data.frame(
        model = model,
        estimand = tis_df$estimand,
        time = tis_df$time,
        state = tis_df$state_set,
        key = tis_df$key,
        estimate = tis_df$estimate,
        std.error = tis_df$std.error,
        p.value = tis_df$p.value
      )
    )
  }
  do.call(rbind, out)
}

replicate_grid <- expand.grid(
  scenario = scenarios$scenario,
  rep = seq_len(N_REP),
  stringsAsFactors = FALSE
)
replicate_grid$plateau <- scenarios$plateau[
  match(replicate_grid$scenario, scenarios$scenario)
]
set.seed(SEED)
replicate_grid$seed <- sample.int(1e9, nrow(replicate_grid))

results <- do.call(rbind, Map(
  function(scenario, plateau, rep, seed) {
    data.frame(
      scenario = scenario,
      rep = rep,
      analyze_replicate(plateau, seed)
    )
  },
  replicate_grid$scenario,
  replicate_grid$plateau,
  replicate_grid$rep,
  replicate_grid$seed
))
```

```{r summarize}
truth_lookup <- rbind(
  data.frame(
    estimand = "sop",
    key = paste(truth_sop_key$time, truth_sop_key$state, sep = "|"),
    scenario = as.character(truth_sop_key$scenario),
    truth = truth_sop_key$truth
  ),
  data.frame(
    estimand = "time_in_state",
    key = paste("tis", truth_key_days$state, sep = "|"),
    scenario = as.character(truth_key_days$scenario),
    truth = truth_key_days$truth
  )
)

results$truth <- truth_lookup$truth[
  match(
    paste(results$estimand, results$key, results$scenario),
    paste(truth_lookup$estimand, truth_lookup$key, truth_lookup$scenario)
  )
]
results$reject <- results$p.value < 0.05
results$scenario <- factor(results$scenario, levels = scenarios$scenario)
results$model <- factor(
  results$model,
  levels = c("constant tx", "tx-by-time")
)

summary_keys <- unique(results[, c("scenario", "model", "estimand", "key")])
performance <- do.call(rbind, lapply(seq_len(nrow(summary_keys)), function(i) {
  rows <- results[
    results$scenario == summary_keys$scenario[i] &
      results$model == summary_keys$model[i] &
      results$estimand == summary_keys$estimand[i] &
      results$key == summary_keys$key[i],
  ]
  power <- mean(rows$reject)
  bias <- mean(rows$estimate) - rows$truth[1]
  data.frame(
    scenario = summary_keys$scenario[i],
    model = summary_keys$model[i],
    estimand = summary_keys$estimand[i],
    key = summary_keys$key[i],
    power = power,
    power_mcse = sqrt(power * (1 - power) / nrow(rows)),
    bias = bias,
    bias_mcse = stats::sd(rows$estimate) / sqrt(nrow(rows)),
    truth = rows$truth[1]
  )
}))
```

## Power for time in state

```{r power-tis}
tis_perf <- performance[performance$estimand == "time_in_state", ]
tis_perf$state <- sub("^tis\\|", "", tis_perf$key)
tis_perf$state <- factor(tis_perf$state, levels = as.character(1:8))
tis_perf$lower <- pmax(0, tis_perf$power - 1.96 * tis_perf$power_mcse)
tis_perf$upper <- pmin(1, tis_perf$power + 1.96 * tis_perf$power_mcse)

power_plot <- function(model_name) {
  ggplot(
    tis_perf[tis_perf$model == model_name, ],
    aes(x = state, y = power, color = scenario, group = scenario)
  ) +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_errorbar(
      aes(ymin = lower, ymax = upper),
      width = 0.2,
      position = position_dodge(width = 0.5)
    ) +
    scale_color_viridis_d() +
    ylim(0, 1) +
    labs(
      x = "State",
      y = "Power (share of p < 0.05) with 95% MCSE interval",
      color = "Scenario",
      title = paste("Time-in-state power:", model_name)
    )
}

power_plot("constant tx")
power_plot("tx-by-time")
```

Power follows the size of the true difference. In the constant scenario,
both models reject in every dataset for the states with large differences
(1 and 4 to 8), and rarely for states 2 and 3, where the true difference in
time in state is essentially zero. As the plateau shortens, the true
differences shrink and so does power: mostly 80–90% with a 15-day plateau,
30–70% with a 10-day plateau, and at most 40% with a 5-day plateau. The two
models have similar power overall; the tx-by-time model estimates more
coefficients, which costs precision, but it also describes the waning effect
better.

The error bars are 95% Monte Carlo intervals, $p \pm 1.96\sqrt{p(1 - p)/R}$.
With $R$ = `r N_REP` they are wide, up to about ±31 percentage points near
50% power. When power is exactly 0 or 1, this formula gives a standard error
of zero, so those error bars have zero width and are invisible.

## Bias in SOP differences and time in state

```{r bias-sop}
sop_perf <- performance[performance$estimand == "sop", ]
sop_perf$time <- as.integer(sub("\\|.*$", "", sop_perf$key))
sop_perf$state <- factor(
  sub("^.*\\|", "", sop_perf$key),
  levels = as.character(1:8)
)
sop_perf$lower <- sop_perf$bias - 1.96 * sop_perf$bias_mcse
sop_perf$upper <- sop_perf$bias + 1.96 * sop_perf$bias_mcse

sop_bias_plot <- function(model_name) {
  ggplot(
    sop_perf[
      sop_perf$model == model_name &
        sop_perf$state %in% c("1", "5", "7", "8"),
    ],
    aes(x = time, y = bias, color = scenario, group = scenario)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point(position = position_dodge(width = 0.8)) +
    geom_errorbar(
      aes(ymin = lower, ymax = upper),
      width = 0.8,
      position = position_dodge(width = 0.8)
    ) +
    facet_wrap(~state, labeller = label_both) +
    scale_color_viridis_d() +
    labs(
      x = "Day",
      y = "Bias in SOP difference with 95% MCSE interval",
      color = "Scenario",
      title = paste("SOP-difference bias:", model_name)
    )
}

sop_bias_plot("constant tx")
sop_bias_plot("tx-by-time")
```

```{r bias-tis}
tis_perf$lower <- tis_perf$bias - 1.96 * tis_perf$bias_mcse
tis_perf$upper <- tis_perf$bias + 1.96 * tis_perf$bias_mcse

tis_bias_plot <- function(model_name) {
  ggplot(
    tis_perf[tis_perf$model == model_name, ],
    aes(x = state, y = bias, color = scenario, group = scenario)
  ) +
    geom_hline(yintercept = 0, linetype = "dashed") +
    geom_point(position = position_dodge(width = 0.5)) +
    geom_errorbar(
      aes(ymin = lower, ymax = upper),
      width = 0.2,
      position = position_dodge(width = 0.5)
    ) +
    scale_color_viridis_d() +
    labs(
      x = "State",
      y = "Bias in time-in-state difference with 95% MCSE interval",
      color = "Scenario",
      title = paste("Time-in-state bias:", model_name)
    )
}

tis_bias_plot("constant tx")
tis_bias_plot("tx-by-time")
```

Because the true values are computed exactly, bias reflects only estimation
error and model misspecification, not simulation noise in the truth. The
first set of plots shows SOP differences for states 1, 5, 7, and 8; the
second shows differences in time in each state.

The bias in daily SOP differences is small: at most about 6 percentage
points, and usually much less, compared with true day-28 differences in
state 1 of 0.04 to 0.11. In the waning scenarios, the constant-tx model tends
to understate the early differences and overstate later ones, because it
spreads the early, full-strength effect over the whole of follow-up. The
tx-by-time model generally follows the decline more closely.

Bias in time in state adds up these daily errors over 28 days, so it can be
substantial. The clearest case is the 5-day plateau: there the constant-tx
model underestimates the gain in days spent in state 1 by about 0.8 days,
against a true gain of 1.5 days, and the tx-by-time model by about 0.5 days.
With longer plateaus, the tx-by-time model's bias in state 1 is close to
zero, whereas the constant-tx model underestimates the gain by roughly 0.3
days. With only `r N_REP` datasets per scenario these estimates are
imprecise, as the wide Monte Carlo intervals show, so a larger simulation
is needed before drawing firm conclusions.